---
title: "Post #13. ggplot Legend Tips Series No.4: Merging ggplot legends"
description: |
  Come and learn two handy tips to merge two ggplot legends into one!
author:
  - name: Gen-Chang Hsu
date: 2022-03-02
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post13.png
draft: TRUE
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## The problem
In this post, we will be using the [`CO2`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/zCO2.html) dataset for our example plots. Let's first create a boxplot of the CO~2~ uptake rates of plant individuals under two chilling treatments. The boxes are colored by the site which the plants are from (column `Type`):   

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)

P1 <- ggplot() + 
  geom_boxplot(data = CO2, aes(x = Treatment, y = uptake, color = Type), width = 0.5, outlier.color = NA) + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

P1
```

Now suppose that we would like to have points on the boxes showing the mean uptake rates. We can compute the means for each `Type` and `Treatment` combination and add them to the boxplot using another geom layer `geom_point()`. Since the column name `Type` doesn't sound intuitive, we'll change it into `Site`:

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
### Mean CO2 uptake rates by Type and Treatment
CO2_mean <- CO2 %>% group_by(Treatment, Type) %>% 
  summarise(uptake_mean = mean(uptake)) %>%
  rename(Site = Type)

### Add the means (different point colors and shapes for the two sites)
P2 <- P1 + geom_point(data = CO2_mean, aes(x = Treatment, y = uptake_mean, color = Site, shape = Site), size = 2.5, position = position_dodge(width = 0.5))

P2
```

Hmm, the two legends (one mapped to "shape" and the other to "color" aesthetic) are not looking good. They shouldn't have been separated! How can we merge them back together? 

## The solutions

Don't panic. It's actually quite simple. There are two ways to fix the problem: 

<span style="display: block; margin-top: 0px; margin-left: 10px"> (1) Rename the legend title and labels</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Override the legend keys</span>

We'll take a look at them shortly below.

### (1) Rename the legend title and labels   

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P2 + scale_color_brewer(name = "Site", palette = "Set1") +
  scale_shape_manual(name = "Site", values = c(16, 17))

```



### Tip 2. Override the legend keys

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P2 + guides(shape = "none", color = guide_legend(title = "Site", override.aes = list(shape = c(16, 17))))

```



## Summary

To recap the two ways to merge ggplot legends, 






Hope you enjoy the reading and don't forget to leave your comments and suggestions below if you have any!



