---
title: "Post #19. Go with the flow - alluvial diagrams in ggplot"
description: |
  Let's create flowy alluvial diagrams in ggplot using the extension package `ggalluvial`!
author:
  - name: Gen-Chang Hsu
date: 2022-09-10
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post19.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

An alluvial diagram is a type of chart that shows the relationships between levels of categorical variables for a given quantity (number of people, amount of energy, etc.) It looks like a stacked barplot with flow streams connecting the stacks of adjacent bars (that's why it gets its name!). Alluvial diagrams are useful for visualizing pathways and trends across variables. Another similar chart is [bump charts](https://genchanghsu.github.io/ggGallery/posts/2022-08-01-post-18-bump-charts-with-ggplot/), which emphasize the change in rankings of multiple objects over time. 

## Create an alluvial plot with `ggalluvial`

Three types of alluvial charts:
1. bars with different levels and colored by another var
2. bars with different levels and colored by one of the bar vars
3. bars with the same levels and colored by another var
4. bars with the same levels and colored by the bar var



```{r, fig.width = 9, fig.height = 5, out.width = "85%"}
# install.packages("ggalluvial")  # install the package if you haven't
library(ggalluvial)
library(tidyverse)





```




## Summary





Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!
