---
title: "Post #25. ggplot Map Series No.1: Creating polygon maps in ggplots"
description: |
  In this post, you will learn about how to create basic polygon maps and choropleth maps in ggplots using the data from the `maps` and `mapdata` package
author:
  - name: Gen-Chang Hsu
date: 2023-03-10
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Map Series"
preview: ../../homepage_images/Post25.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

Welcome to my second blog post series: the ggplot Map Series! Maps are a staple for research, science communication, and even everyday life. In this series, I'll write about making all kinds of maps in ggplots: polygon maps, vector maps, and raster maps, using a variety of data sources from built-in package data sets and online databases, to manually-created data and other external files. Making maps is a critical skill to have under the belt and will certainly go a long way, and I hope this series will come in handy for every reader, you included of course!

## Introduction

There are a diverse set of maps in ggplots. In this first post of the series, I'll just start simple with the most basic type of maps—the polygon maps, using the built-in map data from the `maps` and `mapdata` package.

## Create a polygon map with `maps`

In the following section, we'll first create a basic world map using the world map data in `maps` package. Next, we'll turn this basic map into a choropleth map by adding a theme to it. Finally, we'll take a look at some other available map data in the `maps` package as well as the supplementary `mapdata` package.

### 1. A basic world map

basic world map
projection
background theme

```{r, fig.width = 6, fig.height = 4, out.width = "70%", fig.show = "hold"}
library(tidyverse)
library(maps)



```




### 2. Choropleth map

Add top ten peaks in the world
Add number of bird species by country

```{r, fig.width = 6, fig.height = 4, out.width = "70%", fig.show = "hold"}



```






### 3. Other available map data

```{r, fig.width = 6, fig.height = 4, out.width = "70%", fig.show = "hold"}
library(mapdata)


```







## Summary

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!





