---
title: "Post #25. ggplot Map Series No.1: Creating polygon maps in ggplots"
description: |
  In this post, you will learn about how to create basic polygon maps and choropleth maps in ggplots using the data from the `maps` and `mapdata` package
author:
  - name: Gen-Chang Hsu
date: 2023-03-10
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Map Series"
preview: ../../homepage_images/Post25.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

Welcome to my second blog post series: the ggplot Map Series! Maps are a staple for research, science communication, and even everyday life. In this series, I'll write about making all kinds of maps in ggplots: polygon maps, vector maps, and raster maps, using a variety of data sources from built-in package data sets and online databases, to manually-created data and other external files. Making maps is a critical skill to have under the belt and will certainly go a long way, and I hope this series will come in handy for every reader, you included of course!

## Introduction

There are a diverse set of maps in ggplots. In this first post of the series, I'll just start simple with the most basic type of maps—the polygon maps, using the built-in map data from the `maps` and `mapdata` package.

## Create a polygon map with `maps`

In the following section, we'll first create a world base map using the world map data in `maps` package. Next, we'll turn this base map into a choropleth map by adding a theme to it. Finally, we'll take a look at some other available map data in the `maps` package as well as the supplementary `mapdata` package.

### 1. A basic world map

Let's make a world base map by retrieving the world data from the `maps` package. The returned object is a dataframe with a set of longitude and latitude coordinates that define the border of each country. To draw the map, we'll pass this dataframe into `geom_polygon()` and map the longitude and latitude coordinates to x and y aesthetic. Also, we need to specify the "group" aesthetic so that ggplot knows how the points are grouped in order to form the polygons.

```{r, fig.width = 6, fig.height = 3, out.width = "80%", fig.show = "hold"}
library(tidyverse)
library(maps)

### Get the world map data
world_map <- map_data(map = "world")
head(world_map)

### Make a world base map
P_world_map <- ggplot(data = world_map) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "white", fill = "grey") +  # need to specify the "group" aesthetic
  geom_hline(yintercept = Inf, color = "grey") +  # top panel border
  geom_hline(yintercept = -Inf, color = "grey") +  # bottom panel border
  geom_vline(xintercept = Inf, color = "grey") +  # right panel border
  geom_vline(xintercept = -Inf, color = "grey") +  # left panel border
  theme_void() # empty theme

P_world_map
  
```

Do you notice that there is a "region" column in the world map data containing the country names? The `map_data()` function has a useful argument that allows user to subset the world data by this column. For example, we can get the data for Australia and draw it on the base map using different border and fill colors: 

```{r, fig.width = 6, fig.height = 3, out.width = "80%", fig.show = "hold"}
### Get the Australia data
world_map_Australia <- map_data(map = "world", region = "Australia")

### Draw the Australia map with different colors
P_world_map <- P_world_map + 
  geom_polygon(data = world_map_Australia, aes(x = long, y = lat, group = group), color = "green4", fill = "green3")

P_world_map
  
```

The default map projection is the Mercator projection, in which the true bearing and shape are preserved while the area is increasingly distorted towards the polar regions. We can apply other projections provided by the `mapproj` package. For example, the Mollweide projection sacrifices the bearing and shape for area, and so it is appropriate for maps that require a more accurate representation of country sizes. A full list of the available projections are shown on the help page of `mapproject()`.

```{r, fig.width = 6, fig.height = 3, out.width = "80%", fig.show = "hold"}
library(mapproj)

### Change the map projection to Mollweide projection
P_world_map + 
  coord_map(projection = "mollweide", xlim = c(-180, 180))  # need to set "xlim = c(-180, 180)" to avoid the weird plotting bug 

```

### 2. Choropleth map

Add top ten peaks in the world
Add number of bird species by country

```{r, fig.width = 6, fig.height = 4, out.width = "70%", fig.show = "hold"}



```






### 3. Other available map data

```{r, fig.width = 6, fig.height = 4, out.width = "70%", fig.show = "hold"}
library(mapdata)


```







## Summary

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!





