---
title: "Post #16. ggplot Legend Tips Series No.5: Working on discrete legends with scale_XX functions"
description: |
  In this post, you will learn how to modify the appearance of discrete legends in ggplots using `scale_XX` family of functions.
author:
  - name: Gen-Chang Hsu
date: 2022-06-09
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post16.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

Welcome to the fifth post of **ggplot Legend Tips Series**! This post is a continuation of a [previous post](https://genchanghsu.github.io/ggGallery/posts/2022-01-28-post-10-ggplot-legend-tips-series-no1-working-on-discrete-scale-legends-with-guides/), where I introduced how to modify the appearance of discrete legends in ggplots with `guides()`, such as legend label font, legend key size, and legend direction and position (give it a read if you're interested!).   

Today, we will explore another way to modify discrete legends: using `scale_XX` functions. There are three handy adjustments you can do: 

<span style="display: block; margin-top: 5px; margin-left: 10px"> **(1) Change legend name and labels**</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> **(2) Change the order of legend keys**</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> **(3) Selectively display certain legend keys**</span>

Without further ado, let's jump right in! 

## An example with scale_color_brewer()

Here, we will be using the [`ChickWeight`](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/ChickWeight) dataset for our example plots. The dataset contains the body weights of chicks fed with four different protein diets recorded over a course of 20 days. 

Let's compute the average weight of chicks in each diet treatment and plot the growth curve over time. We'll map diet treatment to the color aesthetic and use `scale_color_brewer()` to set the colors for the legend.

```{r, fig.width = 5.5, fig.height = 3.5, out.width = "70%"}
library(tidyverse)

### Average chick weight over time in each diet treatment
ChickWeight_avg <- ChickWeight %>% 
  group_by(Diet, Time) %>% 
  summarise(weight = mean(weight))

### Growth curve of the chicks
P <- ggplot(ChickWeight_avg) + 
  geom_line(aes(x = Time, y = weight, color = Diet)) + 
  geom_point(aes(x = Time, y = weight, color = Diet)) +  
  labs(x = "Day", y = "Weight (g)") + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

P
```

### (1) Change legend name and labels

Changing legend name and labels is perhaps the most common thing ggplot users would do for their legends. This is super straightforward: simply specify the argument `name = ` for the new legend name and pass a vector of labels to argument `labels = ` in `scale_color_brewer()`. Easy-peasy!!!

```{r, fig.width = 6, fig.height = 3.5, out.width = "75%"}
P + scale_color_brewer(palette = "Set1", 
                       name = "Diet Treatment",  # new legend name
                       labels = c("Red diet", "Blue diet", "Green diet", "Purple diet"))  # new legend labels

```

### (2) Change the order of legend keys

The second thing we can do is reordering the legend keys. As you can see, the chicks fed with "Green" diet seem to grow the best, followed by those fed with "Purple" and "Blue" diet, and those fed with "Red" diet grow the most slowly. So can we change the order of legend keys to reflect this ranking? 

Of course we can! The main argument for this is `break = `, which takes a vector of levels in the legend variable (in this example 1, 2, 3, and 4 in the factor "Diet"). The order of the levels passed to the argument will determine the order of key items displayed in the legend.

```{r, fig.width = 6, fig.height = 3.5, out.width = "75%"}
P + scale_color_brewer(palette = "Set1", 
                       name = "Diet Treatment", 
                       breaks = c(3, 4, 2, 1),  # new order of the legend keys
                       labels = c("Green diet", "Purple diet", "Blue diet", "Red diet"))
```

### (3) Selectively display certain legend keys

Suppose we want to show only the best ("Green") and the worst diet ("Red") in the legend and hide the other two intermediate ones. We can do this by passing the selected levels (here 1 and 3) to the `break = ` argument in a desired order (3 first then 1):

```{r, fig.width = 6, fig.height = 3.5, out.width = "75%"}
P + scale_color_brewer(palette = "Set1", 
                       name = "Diet Treatment", 
                       breaks = c(3, 1),  # select the best and the worst diet
                       labels = c("Green diet (best)", "Red diet (worst)"))


```
<br>

It is worth noting that there is a second way to selectively display certain legend keys, via the argument `limits = `. However, this argument will affect the main plot as well; the levels not specified in `limits =` will not be drawn in the plot! In fact, this is basically the same as dropping the levels in the factor, and the aesthetic mapping will change accordingly. See an example below:

```{r, fig.width = 6, fig.height = 3.5, out.width = "75%"}
P + scale_color_brewer(palette = "Set1", 
                       name = "Diet Treatment", 
                       limits = c(3, 1),  # select the best and the worst diet
                       labels = c("Best diet", "Worst diet"))  # the best and the worst diet are not in green and red now! 

```
<br>

So depending on the purpose of your figure, you might want to use `breaks = ` or `limits = ` to select certain legend keys of interest. But just to keep in mind that the aesthetic mapping will differ between the two methods!

## Summary

In this post, we looked at how to modify a discrete legend via the arguments in `scale_color_brewer()`: `names =` and `labels =` for legend name and labels, as well as `breaks =` and `limits =` for legend keys. The same principle applies to other discrete legend types too, for instance, linetype and shape.

Hope you learn something useful and don't forget to leave your comments and suggestions below if you have any!
