---
title: "Post #23. ggplot Legend Tips Series No.8: The nuts and bolts of legend item spacing"
description: |
  Here are all the nuts and bolts you need for working on legend item spacing!
author:
  - name: Gen-Chang Hsu
date: 2023-01-13
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Legend Tips Series"
preview: ../../homepage_images/Post23.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Welcome to my first blog post in 2023! I have been slacking off a little bit lately and feel so glad to ring in the new year with another [Legend Tips Series](https://genchanghsu.github.io/ggGallery/#category:Legend_Tips_Series) post.

In this post, I will show you how to adjust the spacing between discrete legend items (legend title, keys, labels, etc.) as well as between multiple legends. This is going to be a handy one and let's get to it!

## The nuts and bolts of legend spacing

### 1. Single discrete legend

There are three main arguments in `theme()` when it comes to adjusting the spacing between legend title, legend keys, and legend labels: `legend.title`, `legend.spacing.x`, and `legend.spacing.y`. I'll demonstrate their effects in the following section for vertical legend, horizontal legend, and legend with a matrix layout.

<span style="display: block; margin-top: 0px; margin-left: 0px; font-size: 22px">  (1) Vertical legend</span>

To increase the spacing between legend title and the first legend key, change the bottom margin of the title by specifying `margin = margin(b = )` in the `legend.title` argument. To adjust the horizontal spacing between legend keys and labels, use `legend.spacing.x = unit()` and specify an appropriate value. Similarly, to adjust the vertical spacing between legend keys, use `legend.spacing.y = unit()` and specify an appropriate value. Also, you **must** specify `byrow = T` in the corresponding guides. Otherwise, nothing will happen!

```{r, fig.width = 6, fig.height = 4, out.width = "70%", fig.show = "hold"}
library(tidyverse)

### Default plot
P_vertical_default <- ggplot(data = iris) + 
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_smooth(aes(x = Sepal.Length, y = Sepal.Width, color = Species, fill = after_scale(color)), method = "lm", se = T) + 
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sepal length", y = "Sepal width") + 
  theme_classic(base_size = 13)

### Adjust the spacing between legend items
P_vertical_adjusted <- P_vertical_default + 
  theme(legend.title = element_text(margin = margin(b = 10)),  # change the 
        legend.spacing.x = unit(0.2, "inch"),  # adjust the horizontal spacing between legend keys and labels
        legend.spacing.y = unit(0.2, "inch")) +  # adjust the vertical spacing between legend keys
  guides(color = guide_legend(byrow = T))  # "byrow = T" is necessary for "legend.spacing.y" to work for vertical legends  

P_vertical_default; P_vertical_adjusted

````

(2) Horizontal layout



(3) Matrix layout


### Multiple legends



## Summary

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

















