---
title: "Post #12. ggplot Legend Tips Series No.3: Three ways to turn off the legends in ggplots"
description: |
  Come and learn three handy tips to turn off the legends in ggplots!
author:
  - name: Gen-Chang Hsu
date: 2022-02-05
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post12.png
draft: true
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

There are several things we can modify in the legend:

<span style="display: block; margin-top: 0px; margin-left: 10px"> (1) Legend title</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Legend labels</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (3) Legend keys</span>



### (1)

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)

P_multi_legend <- ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, color = Treatment, shape = Type), position = position_jitter(width = 0.05)) + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

P_multi_legend
```



### (2) 

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}

  
```



### (3)

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}

  
```


## Summary



Hope you enjoy the reading and don't forget to leave your comments and suggestions below if you have any!


