---
title: "Post #12. ggplot Legend Tips Series No.3: Three ways to turn off the legends in ggplots"
description: |
  Come and learn three handy tips to turn off the legends in ggplots!
author:
  - name: Gen-Chang Hsu
date: 2022-02-05
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post12.png
draft: true
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

When we map a variable to an aesthetic (e.g., color), ggplot will automatically create a legend for it. This is quite convenient; however, sometimes the information in the legend may be redundant to the plot itself and we would like to hide the legend. This is actually not a hard task, and in this post I will show you three ways to do so, each for different purposes and with different effects:

<span style="display: block; margin-top: 0px; margin-left: 10px"> (1) Hide the legend for a geom layer</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Hide the legend for an aesthetic mapping</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (3) Hide all the legends in the figure</span>

Let's make an example figure first and we'll walk through the above three ways one by one.

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)

P <- ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, color = Treatment, shape = Type), position = position_jitter(width = 0.05)) + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

P
```

### (1) Hide the legend for a geom layer


```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, color = Treatment, shape = Type), position = position_jitter(width = 0.05), show.legend = F) + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

```



### (2) Hide the legend for an aesthetic mapping 

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P + guides(color = "none")
  
```



### (3) Hide all the legends in the figure

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P + theme(legend.position = "none")
  
```


## Summary



Hope you enjoy the reading and don't forget to leave your comments and suggestions below if you have any!


