---
title: "Post #23. ggplot Legend Tips Series No.8: The nuts and bolts of legend item spacing"
description: |
  Here are all the nuts and bolts you need for working on legend item spacing!
author:
  - name: Gen-Chang Hsu
date: 2023-01-13
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Legend Tips Series"
preview: ../../homepage_images/Post23.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

### Introduction

Welcome to my first blog post in 2023! I have been slacking off a little bit lately and feel so glad to ring in the new year with another [Legend Tips Series](https://genchanghsu.github.io/ggGallery/#category:Legend_Tips_Series) post.

In this post, I will show you how to adjust the spacing between legend items (legend title, keys, labels, etc.) as well as between multiple legends. This is going to be a handy one and let's get to it!

### The nuts and bolts of legend spacing

## 1. Single legend

(1) Vertical layout
(2) Horizontal layout
(3) Matrix layout

```{r, fig.width = 6, fig.height = 6, out.width = "80%"}
library(tidyverse)




````

## Multiple legends



### Summary

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

















