---
title: "Post #20. ggplot Legend Tips Series No.7: Working on legends with theme()"
description: |
    In this post, you will learn how to modify the appearance of ggplot legends using the arguments in `theme()`.
author:
  - name: Gen-Chang Hsu
date: 2022-10-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post20.png
draft: true
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
    
d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;    
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

There are three main groups of functions that work on ggplot legends: `guides()`, `scale_XX` family of functions, and `theme()`. In my previous posts, I introduced how we can modify discrete and continuous legends using the first two function groups (check theme out here! [Post #10](https://genchanghsu.github.io/ggGallery/posts/2022-01-28-post-10-ggplot-legend-tips-series-no1-working-on-discrete-scale-legends-with-guides/), [Post #11](https://genchanghsu.github.io/ggGallery/posts/2022-02-01-post-11-ggplot-legend-tips-series-no2-working-on-continuous-legends-with-guides/), [Post #16](https://genchanghsu.github.io/ggGallery/posts/2022-06-09-post-16-ggplot-legend-tips-series-no5-working-on-discrete-legends-with-scalexx-functions/), and [Post #17](https://genchanghsu.github.io/ggGallery/posts/2022-07-02-post-17-ggplot-legend-tips-series-no6-working-on-continuous-legends-with-scalexx-functions/)). In this post, I will continue to introduce to you the third function "`theme()`". I'll also wrap up this topic by comparing the three function groups and discuss their best usages. Hopefully, you will be able to make the most of these functions to achieve your ggplot goals after the journey! 


## The `theme()` function

`theme()` is one of the main components of ggplots and controls many aspects of the plot appearance: plot title, panel background, axis labels, grid lines, facet strips, and of course legends. There are a bunch of legend-related arguments focusing on different parts of the legends, and here I classify them into four main categories: 

<span style="display: block; margin-top: 0px; margin-left: 10px"> (1) Arguments for legend box</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Arguments for legend keys</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (3) Arguments for legend title and labels</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (4) Arguments for legend layout</span>

Let's first create an example plot with two legends, one discrete and one continuous, using the [`CO2`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/zCO2.html) data set:

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
library(tidyverse)

P <- ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, fill = uptake), position = position_jitter(width = 0.05), shape = 21, size = 2, color = "transparent") + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1") + 
  scale_fill_viridis_c(name = expression(paste(CO[2], " uptake rate")))

P
```

### (1) Arguments for legend box

These arguments control the appearance and margin of the entire legend area as well as individual legend boxes, the arrangement of multiple legends, and the spacing between panel area and legend area: 

```{r, fig.width = 9, fig.height = 5, out.width = "80%"}
P + theme(# the appearance of the entire legend area
          legend.box.background = element_rect(fill = "green1",
                                               color = "black",
                                               size = 1,
                                               linetype = "dashed"),
          
          # the margin of the entire legend area
          legend.box.margin = margin(t = 50, r = 10, b = 50, l = 10),
          
          # the appearance of individual legend boxes
          legend.background = element_rect(fill = "grey90",
                                           color = "black",
                                           size = 0.5,
                                           linetype = "dotted"),
          
          # the margin of individual legend boxes
          legend.margin = margin(t = 5, r = 15, b = 5, l = 15),
          
          # the arrangement of the legends
          legend.box = "horizontal",
          legend.box.just = "right",
          
          # the spacing between panel area and legend area
          legend.box.spacing = unit(0, "inch"))

```

Did you notice that both legends were modified in the same manner? This is an important feature of `theme()`: **it changes everything in the plot (i.e., global effect)**. So if you want to modify just a certain legend, or modify the legends differently (i.e., local effect), you should use `guides()` and specify the specific aesthetic mappings instead.

### (2) Arguments for legend keys

These arguments control the appearance and size (height and width) of legend keys:

```{r, fig.width = 7, fig.height = 5, out.width = "75%"}
P + theme(# the appearance of legend keys
          legend.key = element_rect(fill = "orange",
                                    color = "black",
                                    size = 0.5,
                                    linetype = "solid"),
          
          # the height of legend keys
          legend.key.height = unit(0.3, "inch"),
          
          # the width of legend keys
          legend.key.width = unit(0.6, "inch"))

```

### (3) Arguments for legend title and labels

These arguments control the appearance and alignment of legend title and labels, the spacing between legend keys and labels, and the spacing between legend title and keys/labels:

```{r, fig.width = 7, fig.height = 4, out.width = "75%"}
P + theme(# the appearance of legend title
          legend.title = element_text(size = 15,
                                      margin = margin(l = -10)),
          
          # the alignment of legend title
          legend.title.align = 0.5,
          
          # the appearance of legend labels
          legend.text = element_text(color = "red", 
                                     face = "bold",
                                     angle = 45),
          
          # the alignment of legend labels
          legend.text.align = 0.5,
          
          # the spacing between legend keys and labels
          legend.spacing.x = unit(0.5, "inch"),
          
          # the spacing between legend title and keys/labels
          legend.spacing.y = unit(0.2, "inch"))

```

Here I centered the long title "CO~2~ uptake rate" by specifying a negative left margin to pull the title to the left. Check out my [very first post](https://genchanghsu.github.io/ggGallery/posts/2021-05-01-post-1-center-long-legend-titles-in-ggplots/) on this trick if interested!

### (4) Arguments for legend layout

These arguments control the position of the legend area and the direction of legend items:

```{r, fig.width = 7, fig.height = 5, out.width = "75%"}
P + theme(# the position of the legend area
          legend.position = "top",  # can pass a vector c(x, y) as well
          
          # the direction of legend items
          legend.direction = "horizontal")

```

For the argument `legend.position`, besides the built-in positions ("top", "right", "bottom", "left"), you can also pass a vector of length two `c(x, y)` (between 0 and 1) as the x- and y-coordinate of the legend area relative to the plotting area.

<br>

## Overvall comparison of the three function groups

```{r, echo = FALSE}
library(knitr)
library(kableExtra)

Sum_table <- data.frame(
  "x" = c("General appearance of legend items",
          "Arrangement/layout of legend area/items",
          "Actual text of legend title/labels",
          "Scales (e.g., range) of legend keys",
          "Order of legend keys",
          "Override default mappings",
          "Effect"),
  "guides()" = c("Transformed",
                                    "Equally-spaced",
                                    "Equally-distanced",
                                    "After transformation",
                                    "Not affected",
                 "",
                 ""),
  "scale_XX functions" = c("Original",
                                 "Equally-spaced",
                                 "Not equally-distanced",
                                 "After transformation",
                                 "Not affected",
                           "",
                           ""),
  "theme()" = c("Original",
                                      "Not equally-spaced",
                                      "Equally-distanced",
                                      "Before transformation",
                                      "Might be affected*",
                "",
                "")) %>%
  `colnames<-`(c("", 
                 "guides()", 
                 "scale_XX functions",
                 "theme()"))

kable(Sum_table, align = "ccc") %>%  
  kable_paper(full_width = T, 
              html_font = "Arial",
              font_size = 35) %>%
  row_spec(0, bold = T, color = "black") %>%
  column_spec(1, italic = T)

```

In general, theme and guides have similar effects on legend, focusing more on the physical appearance of the legend items, whereas scales focus more on the actual text (what words should be displayed) and legend scale (the legend range, the legend tick positions, the order of the legend keys). 

There are some specific control for certain methods. For example, guides for override.

Finally, the effects of the funtions differ: theme for global (every legend in the plot) and guides and scales for local (a particular legend).


## Summary

In this post, I showed how you can modify the appearance of ggplot legends using various `theme()` arguments: legend box, legend keys, legend title and labels, and legend layout. I also summarized the three methods for legend modifications and offered some suggestions for when to use them. Of course, the best thing is to have a combination of different methods to achieve the desired outcome! 

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!








