---
title: "Post #20. ggplot Legend Tips Series No.7: Working on legends with theme()"
description: |
    In this post, you will learn how to modify the appearance of ggplot legends using the arguments in `theme()`.
author:
  - name: Gen-Chang Hsu
date: 2022-10-04
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post20.png
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

There are three main legend-related function groups in ggplots: `guides()`, `scale_XX` family of functions, and `theme()`. In my previous posts, I showed how we can modify discrete and continuous legends using the first two function groups (check theme out here! [Post #10](https://genchanghsu.github.io/ggGallery/posts/2022-01-28-post-10-ggplot-legend-tips-series-no1-working-on-discrete-scale-legends-with-guides/), [Post #11](https://genchanghsu.github.io/ggGallery/posts/2022-02-01-post-11-ggplot-legend-tips-series-no2-working-on-continuous-legends-with-guides/), [Post #16](https://genchanghsu.github.io/ggGallery/posts/2022-06-09-post-16-ggplot-legend-tips-series-no5-working-on-discrete-legends-with-scalexx-functions/), and [Post #17](https://genchanghsu.github.io/ggGallery/posts/2022-07-02-post-17-ggplot-legend-tips-series-no6-working-on-continuous-legends-with-scalexx-functions/)). In this post, I will continue to introduce to you the third function "`theme()`". I'll also wrap up this topic by summarizing the features of these three function groups and discuss their usages.  After this journey, I believe you will be able to make the most of these functions to achieve your ggplot goals!


## The `theme()` function

`theme()` is one of the key components of ggplots and controls various aspects of the plot appearance: plot title, panel background, axis labels, grid lines, facet strips, and of course legends. There are quite a few legend-related arguments in `theme()`. To make life easier, I classify them into four main categories based on which parts of the legends they focus on:

<span style="display: block; margin-top: 0px; margin-left: 10px"> (1) Arguments for legend box</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Arguments for legend keys</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (3) Arguments for legend title and labels</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (4) Arguments for legend layout</span>

Let's kick things off by creating an example plot with two legends, one discrete and one continuous, using the [`CO2`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/zCO2.html) data set:

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
library(tidyverse)

P <- ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, fill = uptake), position = position_jitter(width = 0.05), shape = 21, size = 2, color = "transparent") + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1") + 
  scale_fill_viridis_c(name = expression(paste(CO[2], " uptake rate")))

P
```

### (1) Arguments for legend box

These arguments control the appearance and margin of entire legend area as well as individual legend boxes, the arrangement of multiple legends, and the spacing between panel area and legend area: 

```{r, fig.width = 9, fig.height = 5, out.width = "80%"}
P + theme(# the appearance of entire legend area
          legend.box.background = element_rect(fill = "green1",
                                               color = "black",
                                               size = 1,
                                               linetype = "dashed"),
          
          # the margin of entire legend area
          legend.box.margin = margin(t = 50, r = 10, b = 50, l = 10),
          
          # the appearance of individual legend boxes
          legend.background = element_rect(fill = "grey90",
                                           color = "black",
                                           size = 0.5,
                                           linetype = "dotted"),
          
          # the margin of individual legend boxes
          legend.margin = margin(t = 5, r = 15, b = 5, l = 15),
          
          # the arrangement of the legends
          legend.box = "horizontal",
          legend.box.just = "right",
          
          # the spacing between panel area and legend area
          legend.box.spacing = unit(0, "inch"))

```

Did you notice that both legends were modified in the same manner? This is an important feature of `theme()`: **it changes everything in the plot (i.e., global effect)!** So if you want to modify just a certain legend, or modify the legends differently (i.e., local effect), you should use `guides()` instead and specify a specific legend (e.g., the legend for "color" aesthetics) to modify.

### (2) Arguments for legend keys

These arguments control the appearance and size (height and width) of legend keys:

```{r, fig.width = 7, fig.height = 5, out.width = "75%"}
P + theme(# the appearance of legend keys
          legend.key = element_rect(fill = "orange",
                                    color = "black",
                                    size = 0.5,
                                    linetype = "solid"),
          
          # the height of legend keys
          legend.key.height = unit(0.3, "inch"),
          
          # the width of legend keys
          legend.key.width = unit(0.6, "inch"))

```

### (3) Arguments for legend title and labels

These arguments control the appearance and alignment of legend title and labels, the spacing between legend keys and labels, and the spacing between legend title and keys/labels:

```{r, fig.width = 7, fig.height = 4, out.width = "75%"}
P + theme(# the appearance of legend title
          legend.title = element_text(size = 15,
                                      margin = margin(l = -10)),
          
          # the alignment of legend title
          legend.title.align = 0.5,
          
          # the appearance of legend labels
          legend.text = element_text(color = "red", 
                                     face = "bold",
                                     angle = 45),
          
          # the alignment of legend labels
          legend.text.align = 0.5,
          
          # the spacing between legend keys and labels
          legend.spacing.x = unit(0.5, "inch"),
          
          # the spacing between legend title and keys/labels
          legend.spacing.y = unit(0.2, "inch"))

```

In the above figure, I centered the long title "CO~2~ uptake rate" by specifying a negative left margin, which pulled the title to the left. Check out my [very first post](https://genchanghsu.github.io/ggGallery/posts/2021-05-01-post-1-center-long-legend-titles-in-ggplots/) on this trick if interested!

### (4) Arguments for legend layout

These arguments control the position of legend area and the direction of legend items:

```{r, fig.width = 7, fig.height = 5, out.width = "75%"}
P + theme(# the position of legend area
          legend.position = "top",  # can pass a vector c(x, y) as well
          
          # the direction of legend items
          legend.direction = "horizontal")

```

For the argument `legend.position`, besides the built-in positions ("top", "right", "bottom", "left"), you can also place the legend(s) inside the plot by passing a vector of length two `c(x, y)` (between 0 and 1) as the x- and y-coordinate relative to the plot area. For instance, `legend.position = c(0.5, 0.5)` will place the legend(s) in the middle of the plot.


## Comparisons of the three function groups

Here is a summary table of the key features of the three legend-related function groups:

```{r, echo = FALSE}
library(knitr)
library(kableExtra)

Sum_table <- data.frame(
  "x" = c("General appearance of legends",
          "Arrangement/layout of legends",
          "Scales (e.g., range) of legends",
          "Text of legend title/labels",
          "Order of legend keys",
          "Override default legend keys",
          "Effect"),
  "guides()" = c("X",
                 "X",
                 "",
                 "",
                 "",
                 "X",
                 "Local"),
  "scale_XX functions" = c("",
                           "",
                           "X",
                           "X",
                           "X",
                           "",
                           "Local"),
  "theme()" = c("X",
                "X",
                "",
                "",
                "",
                "",
                "Global")) %>%
  `colnames<-`(c("", 
                 "guides()", 
                 "scale_XX functions",
                 "theme()"))

kable(Sum_table, align = "ccc") %>%  
  kable_paper(full_width = T, 
              html_font = "Arial",
              font_size = 35) %>%
  row_spec(0, bold = T, color = "black") %>%
  column_spec(1, italic = T)

```
<br>

In general, `guides()` and `theme()` resemble each other and both control the physical appearance of legends, whereas `scale_XX` functions control the scales (legend range, legend tick positions, etc.) and text (the words displayed in title and labels) of legends. Additionally, there are some specific legend modifications that certain functions can make. For example, `scale_XX` functions can be used to reorder the legend items; `guides()` can be used to overwrite the default aesthetic mappings of legend keys. Finally, these three functions differ in their effects: `theme()` affects every legend in the plot, whereas `guides()` and `scale_XX` functions affect only the specified legend(s).


## Summary

In this post, I showed how you can use various `theme()` arguments to modify the appearance of ggplot legends: legend box, legend keys, legend title and labels, and legend layout. I also summarized the features of three legend-related function groups (`guides()`, `scale_XX`, and `theme()`) and offered some tips for using them. Of course, having a combination of these functions is the best way to achieve the desired outcome!

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!


