---
title: "Post #22. An epic ggplot journey through the glowing Solar System"
description: |
  Ready for an epic ggplot journey through the glowing Solar System? Let's go!   
author:
  - name: Gen-Chang Hsu
date: 2022-11-02
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post22.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Do 
1. Inspired by a recent post by XXX
2. A combination of data scraping, data wrangling, and data visualization techniques
3. Have fun and integrate different skills to produce an interesting figure  

Excited? Take a deep breath and let's get start our space journey!!!


## Create our Solar System step by step

### 1. Scrape the planet data

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
library(tidyverse)
library(rvest)


```


### 2. Crop the planet images

### 3. Create a barplot

### 4. Create a polar barplot

### 5. Polish the plot 


## Summary



Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!
