---
title: "Post #22. An epic ggplot journey through the glowing solar system"
description: |
  Ready for an epic ggplot journey through the glowing solar system? Let's go!   
author:
  - name: Gen-Chang Hsu
date: 2022-11-02
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post22.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

Recently, I stumbled upon a [post](https://www.behance.net/gallery/100683383/Travelling-to-Outer-Space) by data visualization designer Cédric Scherer, and I was completely blown away: I could never imagine how amazing ggplots can be (I believe you'll be in awe of it too)! Inspired by this, I hope to create my own one as well, both for fun and for learning and practicing. In fact, I think this is an excellent opportunity to bring together different data science skills: data scraping, data wrangling, and data visualization.

Now take a deep breath. The rocket is about to lift off and we are ready for our space travel!!!

## Step-by-step guide to the solar system




### Step 1. Scrape planetary data

Scrape the table

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
library(tidyverse)
library(rvest)


```

### Step 2. Add url paths to the planet images

Add wiki url paths to the images

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}



```

### Step 3. Create a line chart of planets

line chart with distance to sun as x axis (reverse the axis) and size as y axis
blurry points with planet sizes as area size, color by dist to sun
add solid opaque points
color the line by distance to sun

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}



```

### Step 4. Create a polar barplot

make the plot circular
add the planet images and labels (white)
add a sun in the middle

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}



```


### Step 5. Polish the figure 

black background
add random dots as stars
add some annotations

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}



```

We're at the very end of the solar system. It's time to get back to Earth. 

## Summary

What a memorable journey! Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!
