---
title: "Post #31. From static to dynamic—Create animations in ggplot by hand"
description: |
  Read below to see how to bring your plots alive!
author:
  - name: Gen-Chang Hsu
date: 2023-12-09
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
preview: ../../homepage_images/Post31.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F,
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

gganimate


But we can make animations in ggplot (or any kind of R plots) by hand. 

Inspired by this post. 

Do some fun coding exercise and write a post on it



## Create an animation in ggplot by hand


### The static plot

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)



```

### Make it alive





## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!






