---
title: "Post #36. The nitty-gritty of dodging in ggplots"
description: |
  Come and take a look at the useful tips for dodging in ggplots!
author:
  - name: Gen-Chang Hsu
date: 2024-08-02
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2024"
preview: ../../homepage_images/Post35.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Dodging is useful when we have two or more groups at the same position and we would like to place them side-by-side to avoid overlapping or to aid comparisons. We can dodge bars, boxes, points, vertical lines, error bars, etc. In this post, I will show you the basics of dodging in ggplots as well as some common dodging issues you might encounter in the wild. Let's get started.

## Basic dodging in ggplots

The basic principle of dodging in ggplots is to specify the group variable by which you would like to dodge in `aes()` and use the function `position_dodge(width = )` to dodge the objects. The variable mapped to the "color", "fill", and "shape" aesthetics will be automatically treated as the group variable for dodging. The distance between the objects can be adjusted via the argument "width". Take a look at the examples below for a dodged barplot, a dodged boxplot, a dodged dotplot, and a dodged line chart.

```{r, fig.width = 9, fig.height = 7, out.width = "90%"}
library(tidyverse)

### (1) A dodged barplot with error bars
ToothGrowth_summary <- ToothGrowth %>% 
  group_by(dose, supp) %>% 
  summarise(mean_len = mean(len),
            n = n(),
            se_len = sd(len)/sqrt(n))

p_dodged_barplot <- ggplot(ToothGrowth_summary) + 
  geom_col(aes(x = as.factor(dose), 
               y = mean_len, 
               fill = supp),  # "supp" will be used as the group variable for dodging
           position = position_dodge(width = 0.7), 
           width = 0.6) + 
  geom_errorbar(aes(x = as.factor(dose), 
                    ymin = mean_len - se_len, 
                    ymax = mean_len + se_len, 
                    group = supp),  # need to specify the group variable for dodging
                position = position_dodge(width = 0.7),  # make sure the dodge width is the same as the previous layer(s) 
                width = 0.1) + 
  scale_fill_brewer(palette = "Dark2") + 
  scale_y_continuous(limits = c(0, 32), expand = c(0, 0)) + 
  labs(x = "Dose", y = "Length", fill = "Supplement") + 
  theme_classic(base_size = 13) + 
  theme(legend.position = c(0.2, 0.85))

### (2) A dodged boxplot with background points
p_dodged_boxplot <- ggplot(ToothGrowth) + 
  geom_boxplot(aes(x = as.factor(dose), 
                   y = len, 
                   color = supp),  # "supp" will be used as the group variable for dodging
               position = position_dodge(width = 0.8), 
               width = 0.6) + 
  geom_point(aes(x = as.factor(dose), 
                 y = len, 
                 color = supp),  # "supp" will be used as the group variable for dodging
             position = position_dodge(width = 0.8), 
             alpha = 0.7) + 
  scale_color_brewer(palette = "Dark2") + 
  labs(x = "Dose", y = "Length", color = "Supplement") + 
  theme_classic(base_size = 13) + 
  theme(legend.position = c(0.2, 0.85))

### (3) A dodged dotchart with line ranges 
p_dodged_dotchart <- ggplot(ToothGrowth) + 
  stat_summary(geom = "pointrange", 
               aes(x = as.factor(dose), 
                   y = len, 
                   color = supp),  # "supp" will be used as the group variable for dodging
               fun.data = "mean_cl_boot",
               position = position_dodge(width = 0.5),
               size = 1.2,
               linewidth = 1) + 
  scale_color_brewer(palette = "Dark2") + 
  labs(x = "Dose", y = "Length", color = "Supplement") + 
  theme_classic(base_size = 13) + 
  theme(legend.position = c(0.2, 0.75),
        legend.key.size = unit(0.4, "in"))

### (4) A dodged line chart
ChickWeight_summary <- ChickWeight %>% 
  group_by(Diet, Time) %>% 
  summarise(mean_weight = mean(weight),
            n = n(),
            se_weight = sd(weight)/sqrt(n))

p_dodged_linechart <- ggplot(ChickWeight_summary) + 
  geom_point(aes(x = Time,
                 y = mean_weight,
                 color = Diet),  # "Diet" will be used as the group variable for dodging
             position = position_dodge(width = 0.4),
             size = 2) + 
  geom_errorbar(aes(x = Time,
                    ymin = mean_weight - se_weight,
                    ymax = mean_weight + se_weight,
                    color = Diet),  # "Diet" will be used as the group variable for dodging
                position = position_dodge(width = 0.4),
                width = 0) + 
  geom_path(aes(x = Time,
                 y = mean_weight,
                 color = Diet),  # "Diet" will be used as the group variable for dodging
            position = position_dodge(width = 0.4)) + 
  scale_color_brewer(palette = "Dark2") + 
  labs(x = "Time", y = "Weight", color = "Diet") + 
  theme_classic(base_size = 13) + 
  theme(legend.position = c(0.2, 0.75),
        legend.key.width = unit(0.5, "in"),
        legend.title = element_text(hjust = 0.5))

### Arrange the plots
library(patchwork)

(p_dodged_barplot + p_dodged_boxplot)/(p_dodged_dotchart + p_dodged_linechart)

```
Now we know the basics, we can move on to some more advanced topics. 

### Handling missing categories







### Dodging with two groups


### Dodging with date/time axis









## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!







