---
title: "Post #35. Creating 3D plots with plotly and rayshader"
description: |
  Bring your plots to the next dimension with `plotly` and `rayshader`!
author:
  - name: Gen-Chang Hsu
date: 2024-07-03
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2024"
preview: ../../homepage_images/Post35.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

2D plots with an x- and a y-axis displaying the relationships between two variables are probably the most common type of plots in our everyday life. If we want to incorporate more information into the plots, we can use different appearances (colors, sizes, shapes, etc.) of points and lines (or other graphic elements). But sometimes, plotting the data in a 3D space can be more visually appealing and may even help us gain deeper insights into the data. Although ggplot does not have geoms for 3D plots by the time of this post, there are other R packages that do the job. And in this post, I'm going to introduce two of these packages to you: `plotly` and `rayshader`. They both have cool functions to generate awesome 3D plots. So buckle up and let's get ready to bring our plots to the next dimension!


## 3D plots with `plotly`

The first package we'll be using is [`plotly`](https://plotly.com/r/). This package features interactive plots and contains functions for generating a wide array of figures. Moreover, besides its own plotting functions, `plotly` also provides the function `ggplotly()`, which directly converts ggplots to the corresponding interactive plots. 

Here, we're going to explore the `plotly` functions for creating 3D plots. Specifically, we'll create a 3D scatterplot with fitted lines and a 3D surface plot with contours and a path.

### (1) 3D scatterplot

The syntax for plotly is pretty similar to ggplot: we call the global function `plotly()` to create a plotting canvas (akin to `ggplot()`) and add other graphical elements using the functions `add_XXX()` (akin to `geom_XXX()`).

To create a basic 3D scatterplot, we use the function `add_markers()` (similar to `geom_point()`). Simply pass the x-, y-, and z-coordinates to the function and specify the variable you would like to color the points by.

To add the lines, we use the function `add_path()` (similar to `geom_path()`). Again, simply pass the x-, y-, and z-coordinates to the function and specify the variable you would like to color the line by.

We can modify the the appearance of the axes and legend using the function `layout()` (see [here](https://plotly.com/r/reference/layout/) for a complete list of things you can adjust!).


```{r, fig.width = 10, fig.height = 5, out.width = "80%"}
library(tidyverse)
library(plotly)

### Model predictions
iris_lm_prediction <- iris %>% 
  group_by(Species) %>% 
  summarise(Model = list(lm(Petal.Length ~ Sepal.Length + Sepal.Width))) %>% 
  rowwise() %>% 
  mutate(Newdata = list(data.frame(Sepal.Length = range(Model$model$Sepal.Length),
                                   Sepal.Width = range(Model$model$Sepal.Width)))) %>% 
  mutate(Prediction = list(cbind(Newdata, Prediction = predict(Model, newdata = Newdata)))) %>% 
  dplyr::select(Species, Prediction) %>% 
  unnest(col = Prediction)

### Create a 3D scatterplot with the fitted lines
scatterplot3d <- plot_ly() %>% 
  # add the points
  add_markers(data = iris, 
              x = ~ Sepal.Length, 
              y = ~ Sepal.Width, 
              z = ~ Petal.Length, 
              color = ~ Species, 
              colors = c("#BF382A", "#0C4B8E", "#018571"),
              size = 2) %>% 
  # add the lines
  add_paths(data = iris_lm_prediction,
            x = ~ Sepal.Length, 
            y = ~ Sepal.Width, 
            z = ~ Prediction,
            color = ~ Species,
            colors = c("#BF382A", "#0C4B8E", "#018571"),
            showlegend = F) %>% 
  layout(legend = list(x = 0.9, y = 0.5),
         scene = list(xaxis = list(title = "Sepal length", range = list(4, 8)),
                      yaxis = list(title = "Sepal width", range = list(1, 5)),
                      zaxis = list(title = "Petal length", range = list(1, 7))))

scatterplot3d

```

### (2) 3D surface plot

The second plot we're going to create is a 3D surface plot with contour lines. This can be done via the function `add_surface()`, which takes a matrix (instead of a dataframe column or a vector) as the input for the argument "z" . The contour lines are added via the argument "contour".

In the example below, we first estimated the kernel density of the Sepal.Length and Sepal.Width in the `iris` dataset using the `kde2d()` from the `MASS` package. The function returns a matrix of the kernel density estimates, with Sepal.Length as the row and Sepal.Width as the column. This matrix can be passed directly to the "z" argument in `add_surface()` to generate a 3D surface plot. Additionally, we can specify the x- and y-axis via the argument "x" and "y". Also note that matrix column is treated as the x-axis (which is "Sepal.Width") and the matrix row is treated as the y-axis (which is "Sepal.Length") in the plot, so make sure you're specifying the correct variables for the axes (it's a bit confusing I know!).

Finally, we can add a random path that traverses across the 3D landscape surface, using the function `add_paths()`.

```{r, fig.width = 10, fig.height = 5, out.width = "80%"}
library(MASS)

### 2D kernel density estimates of the Sepal.Length and Sepal.Width
### The result is a 100-by-100 matrix with Sepal.Length as the row and Sepal.Width as the column
iris_kd_matrix <- kde2d(iris$Sepal.Length, iris$Sepal.Width, n = 100)[[3]]

### Create a 3D surface plot with contours
### Note that the matrix column (Sepal.Width) will be plotted as the x-axis and the matrix row (Sepal.Length) will be plotted as the y-axis
surfaceplot3d <- plot_ly() %>%
  add_surface(x = seq(min(iris$Sepal.Width), max(iris$Sepal.Width), length.out = 100),  # x is the matrix column, corresponding to Sepal.Width
              y = seq(min(iris$Sepal.Length), max(iris$Sepal.Length), length.out = 100),  # y is the matrix row, corresponding to Sepal.Length
              z = iris_kd_matrix,  # takes a matrix as the input
              contours = list(z = list(start = 0, 
                                       end = 0.5, 
                                       size = 0.1,
                                       show = T, 
                                       usecolormap = F))) %>% 
  layout(scene = list(xaxis = list(title = "Sepal width"),
                      yaxis = list(title = "Sepal length"),
                      zaxis = list(title = "Density")))

### Create a dataframe of a random path across the 3D surface
set.seed(123)

random_path_df <- data.frame(x = 1, y = 1)
max_random_path_df <- 1
i <- 1

while (max_random_path_df < 100) {
  random_path_df[i + 1, ] <- random_path_df[i, ] + sample(c(0, 1), size = 2, replace = F)
  max_random_path_df <- max(random_path_df)
  i <- i + 1
}

random_path_df <- random_path_df %>% 
  rowwise() %>% 
  mutate(Density = iris_kd_matrix[x, y]) %>% 
  mutate(Sepal.Length = seq(min(iris$Sepal.Length), max(iris$Sepal.Length), length.out = 100)[x], 
         Sepal.Width = seq(min(iris$Sepal.Width), max(iris$Sepal.Width), length.out = 100)[y])

### Add the random path to the plot
surfaceplot3d <- surfaceplot3d %>% 
  add_paths(data = random_path_df, 
            x = ~ Sepal.Width,  # again x-axis is Sepal.Width 
            y = ~ Sepal.Length,  # again y-axis is Sepal.Length
            z = ~ Density, 
            line = list(width = 4))

surfaceplot3d

```

### (3) Save the plot

To save the plot, we can set the configuration options for it using `config()` and specify the file format, file name, figure width and height (in pixels), and the scale of the figure (a scale larger than 1 will enlarge the figure). After that, adjust the plot to a desired view and click the camera icon "Download plot as a png" on the mode bar to download it.

Alternatively, we can install the application Orca and save the plot directly using the function `orca()`. See [this page](https://plotly.com/r/static-image-export/) for more details. Of course, we can modify the default plot appearance by specifying the x, y, and z values in the "eye" argument before we save it.

```{r, fig.width = 10, fig.height = 5, out.width = "80%"}
### Option 1. Download the plot from the mode bar
### Set the condiguration options and click the download icon
config(surfaceplot3d, toImageButtonOptions = list(format = "png",
                                                  filename = "surfaceplot3d",
                                                  height = 500,
                                                  width = 700,
                                                  scale = 2))

### Option 2. Save the plot using the function orca()
# modify the plot appearance
surfaceplot3d_newview <- surfaceplot3d %>% 
  layout(scene = list(camera = list(eye = list(x = 1.5, y = -1.5, z = 1.5))))

surfaceplot3d_newview

# save the plot
orca(surfaceplot3d_newview, file = "surfaceplot3d.png", width = 700, height = 500, scale = 2)

```

## 3D plots with `rayshader`


```{r, fig.width = 4, fig.height = 3.8, out.width = "80%", fig.show = "hide"}
library(rayshader)

### Create a contour plot in ggplot
contourplot3d <- ggplot(data = iris) +
  geom_density_2d_filled(aes(x = Sepal.Length, y = Sepal.Width, fill = after_stat(nlevel)),
                         n = 100, bins = 20, contour = TRUE) +
  scale_fill_viridis_c() + 
  theme_bw()

### Convert the contour plot into a 3D plot
plot_gg(contourplot3d, 
        width = 3.5, 
        height = 3, 
        shadow_intensity = 0.8,
        scale = 250,
        fov = 25, 
        zoom = 0.6,
        theta = 0,
        windowsize = c(900, 900))

render_snapshot(filename = "contourplot3d")

```


```{r, fig.width = 8, fig.height = 4.5, out.width = "90%", fig.align = "center", echo = FALSE}
ggsave("contourplot3d.tiff", contourplot3d, width = 4.2, height = 3.5, dpi = 600, device = "tiff")

library(cowplot)
ggdraw() + 
  draw_image("contourplot3d.tiff", width = 0.5) + 
  draw_image("contourplot3d.png", width = 0.5, x = 0.5)

```


geom_tile for the iris 2d kde






## Summary






Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!






