---
title: "Post #17. ggplot Legend Tips Series No.6: Working on continuous legends with scale_XX functions"
description: |
  In this post, you will learn how to modify the appearance of continuous legends in ggplots using `scale_XX` family of functions.
author:
  - name: Gen-Chang Hsu
date: 2022-07-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post17.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

This is the part II of the [previous post](https://genchanghsu.github.io/ggGallery/posts/2022-06-09-post-16-ggplot-legend-tips-series-no5-working-on-discrete-legends-with-scalexx-functions/) where I introduced how one can adjust the discrete legends using the `scale_XX` family of functions. In this post, I'm going to talk about adjusting continuous legends, again using the `scale_XX` functions.

## Example


```{r, fig.width = 5.5, fig.height = 3.5, out.width = "70%"}
library(tidyverse)




```














## Summary

In this post, we looked at how to modify a discrete legend via the arguments in `scale_color_brewer()`: `names =` and `labels =` for legend name and labels, as well as `breaks =` and `limits =` for legend keys. The same principle applies to other discrete legend types too, for instance, linetype and shape.

Hope you learn something useful and don't forget to leave your comments and suggestions below if you have any!




