---
title: "Post #18. Bump charts with ggplot"
description: |
  Come and learn how to create nice bump charts with ggplot!
author:
  - name: Gen-Chang Hsu
date: 2022-08-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post18.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Bump charts are a special type of line charts that visualize the rankings of multiple objects over time. These charts are especially useful for comparing and highlighting trends. Honestly, I haven't really heard of this kind of charts before until recently I came across a [Tweet](https://mobile.twitter.com/rappa753/status/1548335653909250048/photo/1) about creating bump charts using ggplot. I then took a deeper dive into it and felt that this would certainly make an interesting topic to blog about. Also, I believe the charts will come in handy some day in the future, both for myself and for the reader like you!

So without further ado, let's jump right in!

## Prepare the data

As mentioned earlier, bump charts are useful for visualizing rankings of items over time. Instead of using a built-in R dataset, this time I'll prepare my own one, which focuses on the rankings of five ecological journals (<span style="color:#0E96D4">*Journal of Ecology*</span>, <span style="color:#7581BD">*Journal of Animal Ecology*</span>, <span style="color:#48A749">*Journal of Applied Ecology*</span>, <span style="color:#E38D26">*Functional Ecology*</span>, and <span style="color:#DE2127">*Methods in Ecology and Evolution*</span>) from the [British Ecological Society](https://www.britishecologicalsociety.org/publications/journals/). Specifically, I would like to see how the rankings of these five journals, in terms of their impact factors, change over the past decade.   

To get the data, I visited [Journal Citation Reports](https://clarivate.com/webofsciencegroup/solutions/journal-citation-reports/) website and downloaded their impact factor records. 

Let's read in the datasets and do some data wrangling.

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}
library(tidyverse)
aa <- read_delim("https://raw.githubusercontent.com/GenChangHSU/ggGallery/master/_posts/2022-08-01-post-18-bump-charts-with-ggplot/BES_journals_data/FUNCT_ECOL.csv", delim = "/")


read_csv("")

```



<br>

## Basic bump chart


```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}

# Use color of the journals
# Func: #e38d26
# Animal: #7581bd
# applied: #48a749
# joe: #0e96d4
# mee: #de2127

```





## A more elaborated bump chart with smooth lines






## Ribbon bump chart







## Summary


Hope you learn something useful from this simple post and don't forget to leave your comments and suggestions below if you have any!









