---
title: "Post #18. Bump charts with ggplot"
description: |
  Come and learn how to create nice bump charts in ggplot!
author:
  - name: Gen-Chang Hsu
date: 2022-08-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post18.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction






## Basic bump chart

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}
library(tidyverse)


```





## A more elaborated bump chart with smooth lines






## Ribbon bump chart







## Summary


Hope you learn something useful from this simple post and don't forget to leave your comments and suggestions below if you have any!









