---
title: "Post #18. Bump charts with ggplot"
description: |
  Come and learn how to create nice bump charts in ggplot!
author:
  - name: Gen-Chang Hsu
date: 2022-08-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post18.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Bump charts are a special type of line charts that plot the rankings of multiple objects over time. These charts are especially useful for comparing and highlighting trends. Honestly, I haven't really heard of this kind of charts before until recently I came across a [Tweet](https://mobile.twitter.com/rappa753/status/1548335653909250048/photo/1) about creating bump charts in ggplot. 

I then took a deeper dive into it and felt that this would make an interesting topic to blog about (also I believe it might come in handy some day in the future)! 

So without further ado, let's jump right in!




## Prepare the data

As mentioned earlier, bump charts are useful for visualizing changes in ranks of items over time. 

Since I study ecology, and in this field, the journals from the British Ecological Society are  

I'm quite interested in how the rankings of these journals, in terms of impact factor, change over the past decade. So I visited JCR and download the impact factor data of these journals (from 1997 to 2021). 

Let's read in the datasets and do some data wrangling


```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}
library(tidyverse)


```



## Basic bump chart


```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}



```





## A more elaborated bump chart with smooth lines






## Ribbon bump chart







## Summary


Hope you learn something useful from this simple post and don't forget to leave your comments and suggestions below if you have any!









