---
title: "Post #38. Creating ternary plots with `ggtern`"
description: |
  Come and learn some useful tips for creating ternary plots in ggplot!
author:
  - name: Gen-Chang Hsu
date: 2024-12-20
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2024"
preview: ../../homepage_images/Post38.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Welcome to the end-of-year post! Really glad to conclude 2024 with this one!

You may (or may not) have heard of it, a [ternary plot](https://en.wikipedia.org/wiki/Ternary_plot#) is a type of plot that displays the relative proportions of three variables in an equilateral triangle. It's a more specialized plot type and certainly not the one you'll see in everyday life, but it can be quite handy when it comes to visualizing compositional data, for example, the carbohydrates, protein, and fat composition of food items.

I remember I first came across a ternary plot in an Ecology textbook years ago when I was learning the classic [Grime's CRS Triangle](https://meristemhorticulture.com/planted/tag/Grime%27s+triangle) for depicting three major plant life strategies (competitor, stress-tolerator, and ruderal) under different stress, resource, and disturbance regimes. I was really fascinated with it, but I didn't know how to make one myself back then. Now I know, and I thought it would be fun (and useful too!) to write a post on creating ternary plots in ggplot. So let's dive in!

## Ternary plots with `ggtern`

The extension package [`ggtern`](https://cran.r-project.org/web/packages/ggtern/index.html) provides a suite of functions for creating ternary plots in ggplot. We'll start from the very basics, gradually add in some variation and complexity, and finally finish by polishing the plot appearance.

### 1. Create a basic ternary plot 

First, let's create a super simple ternary plot with just a single point. 

`ggtern()` is the master function here: it takes a dataframe with three numeric columns, which are then mapped to the "x", "y", and "z" aesthetic. If the three columns are not proportions but instead raw values, the function will automatically rescale them so that each row sums to 1.

In the example below, I also drew the lines connecting the point to the sides using the function `geom_crosshair_tern()` and added the arrows and labels along the axes using the function `theme_arrownormal()`. These will help us better understand what the plot is doing.

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)
library(ggtern)

### Generate the data
ternary_df_point <- data.frame(x = 1, y = 2, z = 3)

### Create a basic ternary plot
ggtern(data = ternary_df_point, aes(x = x, y = y, z = z)) +  # the values will be rescaled 
  geom_point() +
  geom_crosshair_tern() +  # draw the lines
  theme_arrownormal()  # add the arrows and labels

```

If you're new to ternary plots, you may be scratching your head about how to read the plot. Well, it's not that difficult once you understand the principle: as indicated by the lines, the point corresponds to an x-value of around 17% (1/6), a y-value of around 33% (2/6), and a z-value of 50% (3/6). 

For plots that do not have the lines (which are most likely what you'll see in the wild), you can draw mental lines from the point to the sides with the same angle as the numbers. So for example, the numbers on the "x" side have an angle of 120°, so you can draw a line with an angle of 120° from the point to the side and see what the corresponding value is.

### 2. From points to lines

Now that we have a better idea of how ternary plots work, we can take a step further and add . Instead of just showing the individual points, we can actually connect them to create a "ternary line chart". This will be useful for visualizing how compositions change over time.

```{r, fig.width = 5, fig.height = 4}
### Generate the data
ternary_df_line <- data.frame(x = seq(2, 9, length.out = 10)^2, 
                              y = seq(9, 2, length.out = 10)^2) %>% 
  mutate(z = 100 - x - y) %>% 
  rowid_to_column(var = "time")

### Create a ternary line chart
ggtern(data = ternary_df_line, aes(x = x, y = y, z = z)) +
  geom_point(aes(color = time)) +
  geom_line(aes(color = time)) +
  scale_color_viridis_c(breaks = c(1, 10)) + 
  guides(color = guide_colorbar(reverse = T)) + 
  theme_arrownormal()

```

### 3. Contour ternary plot

Next, we'll go from lines to areas by adding density contours to the ternary plot.

Remember I mentioned that one application of ternary plots is to visualize the nutrient composition of food, and this is what we're going to do now.

We'll show the carbs, protein, and fat proportion of different food items at starbucks using the [`starbucks`]() dataset from the package `openintro`.

We can add the contours using the function `stat_density_tern()`. We also used `theme_rgbw()` to color the sides for easier reading. 


```{r, fig.width = 5, fig.height = 4}
library(openintro)  # for the dataset "starbucks"

data("starbucks")

ggtern(data = starbucks,
       mapping = aes(x = carb, y = protein, z = fat)) +
  stat_density_tern(geom="polygon",
                     aes(fill  = after_stat(nlevel),
                         alpha = after_stat(nlevel)),
                    binwidth = 0.5) + 
      stat_density_tern(geom="polygon",
                      coutour = T,
                      color = "black",
                      fill = NA,
                    binwidth = 5,
                    size = 0.05) + 
    geom_point(size = 1, alpha = 0.7) +
    scale_fill_distiller(name = "Relative density", palette = "Reds", direction = 1,
                         breaks = c(0.1, 0.3, 0.5, 0.7, 0.9)) + 
  guides(alpha = "none") + 
    theme_rgbw() + 
  theme(legend.position = c(1.05, 0.5),
        legend.title = element_text(hjust = 0.5),
        plot.margin = margin(r = 70))

```
Seems like most food at Starbucks is high in carbs and low in protein (not surprising?!).




### 4. hex plot and tri plot



### 5. Interpolation of z values



### 6. Adjust the appearance and axis and annotations













## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!