---
title: "Post #10. ggplot Legend Tips Series No.1: Working on discrete legends with guides()"
description: |
  In this post, you will learn how to tweak the appearance of discrete legends in ggplots with `guides()`.
author:
  - name: Gen-Chang Hsu
date: 2022-01-28
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post10.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

Welcome to the first post of 2022—also my 10^th^ ggGallery post! A tiny but meaningful milestone reached!

## Background

Legends are arguably one of the greatest advantages of ggplots over the R base graphics; they are so versatile that you can basically create anything you want. Though, it often takes quite a while to wrap your head around lots of argument details. Also, there are several "players" controlling the legends in ggplots—`scale`, `guide`, and `theme`, adding further complications to the already-complicated system.

Inspired by the problems I've personally encountered throughout my ggplot learning journey and the questions my friends have asked me about, I've decided to launch a blog post series "Legend Tips", where I dive into the relevant functions/arguments and show how you can make use them to enhance your ggplot legends, as well as provide some handy tips for dealing with frequently-encountered legend issues. I feel that legends would be a great topic to write about, both for myself and for those geeks who use/love ggplots so much as I do! 

We'll kick this series off by working on discrete legends using the function `guides()`. Keep reading!


## Discrete legends

There are two main types of legends in ggplots: **discrete** and **continuous**. Discrete legends correspond to the aesthetics mapped to categorical variables (factors), for example, gender and eye color; continuous legends correspond to the aesthetics mapped to continuous variables, for example, tree height and population density. These two legend types can live together in the same figure. They do have many commonalities, but still differ in some detail specifications.

In this post, we will be using the [`CO2`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/zCO2.html) dataset, which contains the CO~2~ uptake rates of 12 grass plant individuals ([_Echinochloa crus-galli_](https://en.wikipedia.org/wiki/Echinochloa_crus-galli)) from two sites Quebec and Mississippi (each 6) measured at several CO~2~ concentrations. Additionally, half of the plant individuals from each site were chilled overnight before measurement.

Let's first take a look at the CO~2~ uptake rates under the two treatments (nonchilled vs. chilled):

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)

### Boxplot of CO2 uptake rates by treatment
P <- ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, color = Treatment), position = position_jitter(width = 0.05)) + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

P
```

Seems that nonchilled individuals have higher CO~2~ uptake rates than chilled ones by visual inspection. Well, this is not the focus of this post. Let's get to the main topic and see how we can modify the legend using `guides()`. 


## The `guide_legend()` function

The actual function in `guides()` that controls the discrete legends is `guide_legend()`. In this example, we mapped the factor "Treatment" to color, so we will pass `color = guide_legend()` into `guides()` and specify the arguments in `guide_legend()`.

There are several things we can modify in the legend:

<span style="display: block; margin-top: 0px; margin-left: 10px"> (1) Legend title</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Legend labels</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (3) Legend keys</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (4) Legend layout</span>

Additionally, we can (5) change the default aesthetic mappings and (6) make specific adjustments for multiple legends.

### (1) Legend title

We can modify the name, position (relative to the legend labels and keys), and text appearance of the title. We can also adjust its horizontal and vertical alignment.

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P + guides(color = guide_legend(title = "Treatment",  # name
                                title.position = "top",  # position
                                title.theme = element_text(size = 15, color = "darkgreen", face = "bold.italic"),  # text appearance
                                title.hjust = 0.5,  # horizontal alignment
                                title.vjust = 2))  # vertical alignment
  
```

### (2) Legend labels

Similar to what we've done for the title, we can modify the position (relative to the legend keys) and text appearance of the labels as well as adjust the horizontal and vertical alignment. If we want to hide the labels, specify `label = F`.

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P + guides(color = guide_legend(label = T,  # show the labels
                                label.position = "top",  # position
                                label.theme = element_text(size = 12, color = "brown", face = "italic"),  # text appearance
                                label.hjust = 0.5,  # horizontal alignment
                                label.vjust = 0.5))  # vertical alignment
  
```

### (3) Legend keys

We can change the width and height of the legend keys. This is especially handy for adjusting the length (width) of line segments.

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P + guides(color = guide_legend(keywidth = unit(0.25, "inches"),
                                keyheight = unit(0.5, "inches")))
  
```

### (4) Legend layout

We can display the legend either horizontally or vertically, with legend items separated into several rows and columns. We can also reverse the order of the items.

```{r, fig.width = 7.5, fig.height = 3.5, out.width = "85%"}
P + guides(color = guide_legend(direction = "horizontal",  # direction of the legend
                                nrow = 1,  # number of rows
                                ncol = 2,  # number of columns
                                byrow = T,  # arrange the legend items row by row 
                                reverse = T))  # reverse the legend items
  
```

### (5) Change default mappings

Suppose that we want to change the default aesthetic mappings, for example, the color of boxes, the shape of the points, and the size of the legend keys. We can do this by passing a list of new aesthetics to the argument `override.aes`:

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P + guides(color = guide_legend(override.aes = list(color = c("purple", "green"),
                                                    shape = 17,
                                                    size = 1)))
  
```

### (6) Multiple legends

Recall that the plant individuals come from two sites Quebec and Mississippi (variable "Type" in the dataset). We can map it to the "shape" aesthetic:

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
### Different shapes for the two sites
P_multi_legend <- ggplot(data = CO2) + 
  geom_boxplot(aes(x = Treatment, y = uptake, color = Treatment), width = 0.5) + 
  geom_point(aes(x = Treatment, y = uptake, color = Treatment, shape = Type), position = position_jitter(width = 0.05)) + 
  labs(x = "Treatment", y = expression(paste(CO[2], " uptake rate"))) + 
  theme_classic(base_size = 14) + 
  scale_color_brewer(palette = "Set1")

P_multi_legend
```

No we have two legends in the figure; time to make some changes for them!

Here, we apply `guide_legend()` individually to color and shape aesthetic and specify the respective arguments. Also, we use an additional argument `order` to tell ggplot the order in which the legends should be placed. 

```{r, fig.width = 6, fig.height = 4, out.width = "70%"}
P_multi_legend + guides(color = guide_legend(title.position = "top",
                                             title.hjust = 0.5,
                                             label.position = "top",
                                             keywidth = unit(0.7, "inches"),
                                             keyheight = unit(0.8, "inches"),
                                             direction = "horizontal",
                                             order = 1),
                        shape = guide_legend(title = "Site",
                                             title.hjust = 0.5,
                                             label.position = "left",
                                             label.hjust = 0.5,
                                             label.vjsut = 0.5,
                                             keywidth = unit(0.5, "inches"),
                                             reverse = T,
                                             override.aes = list(size = 3),
                                             order = 2)
                        )
```
The changes don't really make sense to me. Of course, it's just a toy example showing the principles so that you can apply them next time to real-life situations.

Well that's pretty much for now. But before we finish, it is worth mentioning that many of the arguments in `guides()` actually overlap with those in another legend function `theme()` (which will be the topic in another post of this series). So what's the difference between these two functions? Which one should you use?

So the main difference is that, the arguments in `guides()` only affect the legends corresponding to the specified aesthetics, whereas the arguments in `theme()` will affect all the legends in the figure together. So if you have only one legend, using either function would give the same results. However, if you have multiple legends and you want to make different adjustments for each of them, then `guides()` is the go-to.


## Summary

In this post, we've learned how to modify the appearance and layout of legend items in discrete ggplot legends using the function `guides()`. We can even customize the legend keys by overriding the default aesthetic mappings. Try playing around with the arguments a bit and I believe you will be able to make a nice legend(s) for your figure!

Hope you enjoy the reading and don't forget to leave your comments and suggestions below if you have any!


