---
title: "Post #10. ggplot Legend Tips Series No.1: Working on discrete-scale legends with guides()"
description: |
  In this post, you will learn how to tweak the appearance of discrete-scale legends in ggplots with `guides()`.
author:
  - name: Gen-Chang Hsu
date: 2022-01-28
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post10.png
draft: true
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

<!-- Start of the article -->

Welcome to the first post in 2022—also my 10^th^ ggGallery post! A tiny but meaningful milestone reached!

In this new year, I am going to launch a blog post series "Legend Tips", where I dive into the relevant functions/arguments to show how you can make use them to enhance your ggplot legends, as well as provide some handy tips for dealing with some commonly-encountered legend issues. 


## Introduction

Legends are arguably one of the greatest advantages of ggplots over the R base graphics; ggplot legends are so versatile that you can basically do anything you want. However, there is also a downside of this advantage: it often takes quite a while to wrap your head around lots of argument details. What's worse, there are several different "systems" controlling legends in ggplots—scales, guides, and themes, adding further complications to the already-complicated    

Inspired by the legend problems I've personally encountered throughout my ggplot learning journey as well as the questions my friends have asked me about, I feel that legends would be a great topic to write about, both for myself and for any geeks who love ggplots so much as I do!

Let's kick this series off with working on discrete-scale legends using the function `guides()`.


## Discrete-scale legends

We can modify the legends mapped to a continuous variable using `guide_legend()`

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)

```



## Summary

In this post, we've seen how to use the arguments in `guides()` to work on the discrete-scale legends. 

We can modify the appearance of legend keys, layouts, positions. We can even customizing the legend by overriding the default aesthetics.


Hope you enjoy the reading and don't forget to leave your comments and suggestions below if you have any!


