---
title: "Post #17. ggplot Legend Tips Series No.6: Working on continuous legends with scale_XX functions"
description: |
  In this post, you will learn how to modify the appearance of continuous legends in ggplots using `scale_XX` family of functions.
author:
  - name: Gen-Chang Hsu
date: 2022-07-03
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
  - "Legend Tips Series"
preview: ../../homepage_images/Post17.png
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

This is the part II of the [previous post](https://genchanghsu.github.io/ggGallery/posts/2022-06-09-post-16-ggplot-legend-tips-series-no5-working-on-discrete-legends-with-scalexx-functions/) where I introduced how one can adjust the discrete legends in ggplots using the `scale_XX` family of functions. In this post, I'm going to talk about adjusting continuous legends, again using the `scale_XX` functions.

## An example with scale_fill_gradient()

Let's start with an example plot of eruption time vs. waiting time of the old faithful geyser using the [`faithful`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/faithful.html) dataset. In the below code chunk, the function `geom_hex()` divides the plot panel into a grid of hexagons, counts the number of original data points in each hexagon, and maps the counts to the "fill" aesthetic. This is a useful way to avoid overlapping points (i.e., overplotting), especially when the dataset is huge.

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}
library(tidyverse)

ggplot(faithful, aes(x = eruptions, y = waiting)) + 
  geom_hex(aes(fill = ..count..)) +  # hexagon bins 
  labs(x = "Eruption time (min)", y = "Waiting time (min)") + 
  theme_classic(base_size = 13)

```

<br>

Now we can try to adjust the legend bar using the arguments in `scale_fill_gradient()`. Three main things to change are:

<span style="display: block; margin-top: 5px; margin-left: 10px"> (1) Legend range, controlled by the argument `limits`</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (2) Legend tick positions, controlled by the argument `breaks` </span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> (3) Legend tick labels, controlled by the argument `labels` </span>

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}
ggplot(faithful, aes(x = eruptions, y = waiting)) + 
  geom_hex(aes(fill = ..count..)) +
  scale_fill_gradient(name = "Number of \n eruptions",
                      limits = c(0, 6),  # legend range
                      breaks = c(1, 2, 3, 4, 5),  # legend tick positions
                      labels = format(c(1, 2, 3, 4, 5), nsmall = 1)) +  # legend tick labels
  labs(x = "Eruption time (min)", y = "Waiting time (min)") + 
  theme_classic(base_size = 13)

```

<br>

We can also cut the legend bar into several bins by using the argument `guide = guide_colorsteps()`:

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%"}
ggplot(faithful, aes(x = eruptions, y = waiting)) + 
  geom_hex(aes(fill = ..count..)) +
  scale_fill_gradient(name = "Number of \n eruptions",
                      limits = c(0, 6),  
                      breaks = c(1, 2, 3, 4, 5),  
                      labels = format(c(1, 2, 3, 4, 5), nsmall = 1),
                      guide = guide_colorsteps()) +  # cut the legend bar into bins
  labs(x = "Eruption time (min)", y = "Waiting time (min)") + 
  theme_classic(base_size = 13)

```

## Summary

In this post, we've seen how to adjust a continuous legend via the arguments in `scale_fill_gradient()`: `limits` for setting the range of the legend bar, `breaks` for drawing the legend ticks, and `labels` for adding the tick labels. We can also convert the continuous legend bar into a discrete one by specifying `guide = guide_colorsteps()`. 

Hope you learn something useful from this simple post and don't forget to leave your comments and suggestions below if you have any!



