---
title: "Post #41. ggplot Map Series No.7: Spatial raster mapping in ggplot"
description: |
  In this post, you will learn how to create maps in ggplot using spatial raster data!
author:
  - name: Gen-Chang Hsu
date: 2025-05-05
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2025"
  - "Map Series"
preview: ../../homepage_images/Post41.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

What are spatial raster data?






## Spatial raster mapping in ggplot

We'll use the [WorldClim 1970-2000 5-km^2^ global monthly precipitation data](https://www.worldclim.org/data/worldclim21.html), which contain 12 GeoTIFF files for each month. These GeoTIFF files can be read into R via the function `rast()` from the package `terra` as "SpatRaster" objects.

```{r}
library(tidyverse)
library(terra)

### Download the WorldClim data
download.file("https://geodata.ucdavis.edu/climate/worldclim/2_1/base/wc2.1_2.5m_prec.zip",
              destfile = "./WorldClim_Precipitation.zip")
unzip("WorldClim_Precipitation.zip", exdir = "WorldClim")

### Read the GeoTIFF files
worldclim_filepath <- list.files("./WorldClim", pattern = ".tif") %>% 
  str_c("./WorldClim/", .)
worldclim_precipitation_raster <- map(worldclim_filepath, rast)  # rast() reads the GeoTIFF files as "SpatRaster" objects

```

### Basic visualization



```{r, fig.width = 5, fig.height = 4}
library(tidyverse)



```


### Add country boundaries





### Change map projections





### Create a chloropheth map using country average





### Create an animation of monthly average






## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

