---
title: "Post #41. ggplot Map Series No.7: Spatial raster mapping in ggplot"
description: |
  In this post, you will learn how to create maps in ggplot using spatial raster data!
author:
  - name: Gen-Chang Hsu
date: 2025-05-05
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2025"
  - "Map Series"
preview: ../../homepage_images/Post41.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

What are spatial raster data?






## Spatial raster mapping in ggplot

### 1. Data preparation

We'll use the [WorldClim 1970-2000 5-km^2^ global monthly precipitation data](https://www.worldclim.org/data/worldclim21.html), which contain 12 GeoTIFF files for each month. These GeoTIFF files can be read into R via the function `rast()` from the package `terra` as "SpatRaster" objects.

```{r cache = T}
library(tidyverse)
library(terra)

### Download the WorldClim data
download.file("https://geodata.ucdavis.edu/climate/worldclim/2_1/base/wc2.1_2.5m_prec.zip",
              destfile = "./WorldClim_Precipitation.zip")
unzip("WorldClim_Precipitation.zip", exdir = "WorldClim")

### Read the GeoTIFF files
worldclim_filepath <- list.files("./WorldClim", pattern = ".tif") %>% 
  str_c("./WorldClim/", .)
worldclim_precipitation_raster <- map(worldclim_filepath, rast)  # rast() reads the GeoTIFF files as "SpatRaster" objects

```

### 2. Create a basic raster map

The object "worldclim_precipitation_raster" contains 12 lists of SpatRaster objects, each of which has x- and y-coordinates as well as a layer for the precipitation. Sometimes, a SpatRaster object may have more than one layer, and we can get all the layer names using `names()` from the package `tidyterra`. 

Let's make a map of the average global precipitation in January (the first list in "worldclim_precipitation_raster"). To create a map from a SpatRaster object, we can use the function `geom_spatraster()` from the package `tidyterra`. This function is a wrapper for `geom_raster()`. In this example, we only have one layer, so it's not necessarily to specify the layer for the "fill" aesthetic, but we need to do so if there are multiple layers. We'll also draw the country boundaries from the package `rnaturalearth` on the map.

```{r, out.width = "80%", fig.width = 8, fig.height = 6, cache = T}
### Get country boundaries
library(rnaturalearth)
country_boundary <- ne_countries(type = "countries", scale = "small")

### Create a palette for the precipitation
library(colorspace)
col_pal <- sequential_hcl(n = 7, palette = "BluYl", rev = T)

### Get raster layer names
library(tidyterra)
names(worldclim_precipitation_raster[[1]])

### Create a map
ggplot() +
  geom_spatraster(data = worldclim_precipitation_raster[[1]]) +  # January precipitation
  geom_sf(data = country_boundary, fill = NA, color = "black", size = 0.8) +  # country boundaries
  scale_fill_gradientn(colors = col_pal[c(1, 4, 6, 7)], 
                       na.value = "transparent",
                       name = "January precipitation (mm)") + 
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = F) + 
  theme_bw(base_size = 13) +  
  theme(legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(0.8, "in"),
        legend.title = element_text(hjust = -5, vjust = 0.75, size = 12),
        legend.text = element_text(size = 10))

```

### Change the map projection



```{r}
### Transform the map projection to Robinson
terraOptions(progress = 0)
january_precipitation_raster_proj <- project(worldclim_precipitation_raster[[1]], "ESRI:54030")

### Create the long-lat lines and axis labels
library(sf)
country_boundary_proj <- st_transform(country_boundary, crs = "ESRI:54030")


grid <- list(
       st_sf(geometry = st_sfc(
  st_multilinestring(x = lapply(c(-180, -120, -60, 0, 60, 120, 180), 
                function(x) cbind(x, seq(-90, 90, 1)))), crs = 'WGS84')),
        st_sf(geometry = st_sfc(
  st_multilinestring(x = lapply(c(-90, -60, -30, 0, 30, 60, 90), function(x) {
    cbind(seq(-180, 180, 1), x)
  })), crs = 'WGS84'))) %>% bind_rows()

ylabs <- lapply(c(-90, -60, -30, 0, 30, 60, 90), function(x) {
  st_sf(label = paste0(abs(x), '\u00b0', 
                       ifelse(x == 0, '', ifelse(x < 0, 'S', 'N'))),
        geometry = st_sfc(st_point(c(-180, x)), crs = 'WGS84'))
}) %>% bind_rows()

### Make the map
ggplot() +
  geom_spatraster(data = january_precipitation_raster_proj) +  # January precipitation
  geom_sf(data = country_boundary_proj, fill = NA, color = "black", size = 0.8) + 
    geom_sf(data = grid, linewidth = 0.1, color = "grey") +
  geom_sf_text(data = ylabs, aes(label = label), size = 3, color = 'gray30',
               nudge_x = c(-2e6, -1.5e6, -1.3e6, -1e6, -1.3e6, -1.5e6, -2e6)) +
# country boundaries
  # scale_x_continuous(expand = c(0, 0)) +
  # scale_y_continuous(expand = c(0, 0)) +
  scale_fill_gradientn(colors = col_pal[c(1, 4, 6, 7)], 
                       na.value = "transparent",
                       name = "January precipitation (mm)") + 
  theme_minimal() +
    coord_sf(label_axes = "E--N",
           expand = FALSE,
           crs =  "ESRI:54030",
           clip = "off") +
  theme(legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(0.8, "in"),
        legend.title = element_text(hjust = -5, vjust = 0.75, size = 12),
        legend.text = element_text(size = 10),
        panel.border = element_blank(),
        plot.background = element_blank(),
                panel.grid = element_blank(),
        axis.title = element_blank())

```





### Create a chloropheth map using country average





### Create an animation of monthly average






## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

