---
title: "Post #27. ggplot Map Series No.3: Visualizing the veins of the earth in ggplots with `osmdata`"
description: |
  Create a map of the rivers in the world in ggplots using the package `osmdata`.
author:
  - name: Gen-Chang Hsu
date: 2023-7-31
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Map Series"
preview: ../../homepage_images/Post27.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

A brief introduction of the OSM data





## Create the map



### Prepare the data

```{r, fig.width = 6, fig.height = 3, out.width = "80%"}
library(tidyverse)
library(osmdata)






```


### Create a basic map

Make a base world map using the maps package
Draw the rivers using the geom_sf() function

```{r, fig.width = 6, fig.height = 3, out.width = "80%"}
library(maps)






```


### Make it artistic

Color the river by continent
Set the transparency by river length

```{r, fig.width = 6, fig.height = 3, out.width = "80%"}






```



## Summary



Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

