---
title: "Post #43. Pair them up—creating scatterplot matrices and correlation heatmaps in ggplots"
description: |
  In this post, you will learn how to create awesome scatterplot matrices and correlation heatmaps in ggplots using the packages `GGally` and `ggcorheatmap`!
author:
  - name: Gen-Chang Hsu
date: 2026-02-17
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2026"
preview: ../../homepage_images/Post43.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Welcome to the first post of the year 2026! It's been over a half year since my last post, so supper happy to be back in the saddle!

Suppose you have a dataset with several variables in the columns (can be continuous, categorical, or a mix of both) and you would like to explore the bivariate relationships between the variables, then a scatterplot matrix will be really handy for this: it visualizes all pairwise combinations of variables using a range of plot types, for example, a scatterplot for two continuous variables and a boxplot for one continuous vs. one categorical variable. If you have just continuous variables in the data, you can further visualize their pairwise correlation coefficients in a correlation heatmap, which uses color for the sign of correlations and intensity for the strength of correlations. Interesting patterns will often reveal themselves!

In this post, we'll be creating scatterplot matrices in ggplots using the extension package `GGally` as well as correlation matrices/heatmaps using the extension package `ggcorrheatmap`. Let's start pairing things up!

## Part I. Create scatterplot matrices with `GGally`

### (1) Basic scatterplot matrices

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)
library(GGally)




```





## Part II. Create correlation heatmaps with `ggcorrheatmap`
`ggcorrheatmap` offers a variety of functionality to create basic and more customized correlation heatmaps. Let's start with the basic ones.

### (1) 


```{r, fig.width = 5, fig.height = 4}
library(ggcorrheatmap)

### Basic correlation heatmap
# change cell color
# change cell shape
# change the labels




```


### (2)

```{r, fig.width = 5, fig.height = 4}
library(ggcorrheatmap)

### Basic correlation heatmap
# change cell color
# change cell shape
# change the labels


### Mixed layout
# change layout
# tile + p-values + significance stars


### Modify cell appearance
# highlight cell border


### Cluster the variables and add annotation






```




## Summary




Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

