---
title: "Post #43. Pair them up—creating scatterplot matrices and correlation heatmaps in ggplots"
description: |
  In this post, you will learn how to create awesome scatterplot matrices and correlation heatmaps in ggplots using the packages `GGally` and `ggcorheatmap`!
author:
  - name: Gen-Chang Hsu
date: 2026-02-17
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2026"
preview: ../../homepage_images/Post43.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Welcome to the first post of the year 2026! It's been over a half year since my last post, so supper happy to be back in the saddle!

Suppose you have a dataset with several variables in the columns (can be continuous, categorical, or a mix of both) and you would like to explore the bivariate relationships between the variables, then a scatterplot matrix will be really handy for this: it visualizes all pairwise combinations of variables using a range of plot types, for example, a scatterplot for two continuous variables and a boxplot for one continuous vs. one categorical variable. If you have just continuous variables in the data, you can further visualize their pairwise correlation coefficients in a correlation heatmap, which uses color for the sign of correlations and intensity for the strength of correlations. Interesting patterns will often reveal themselves!

In this post, we'll be creating scatterplot matrices in ggplots using the extension package `GGally` as well as correlation matrices/heatmaps using the extension package `ggcorrheatmap`. Let's start pairing things up!

## Part I. Create scatterplot matrices with `GGally`

### (1) Basic scatterplot matrices

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)
library(GGally)
library(carData)

data("Soils")


```





## Part II. Create correlation heatmaps with `ggcorrheatmap`

In the second part of this post, we'll focus on the continuous variables in the `Soil` dataset. `ggcorrheatmap` offers a wide variety of functionality for creating correlation heatmaps. Let's start with the basics and add more complexity as we go.

### (1) A basic correlation heatmap
`ggcorrhm()` is the master function for creating correlation heatmaps. Simply pass your data to the function and you'll get a basic heatmap (note that he default choice of correlation is the Pearson correlation coefficient)!. We can also adjust the cell color and shape, and remove the diagonal cells (which all have a correlation coefficient of 1).

```{r, fig.width = 5, fig.height = 4}
library(ggcorrheatmap)

### Select continuous variables
Soils_continuous <- Soils %>% 
  select(is.numeric)

### Create a basic correlation heatmap
ggcorrhm(Soils_continuous)

### Change the cell color and shape
ggcorrhm(Soils_continuous,
         high = "#de2d26",  # the color for the upper limit
         mid = "white",  # the color for the midpoint
         low = "#006d2c",  # the color for the lower limit
         mode = 21,  # the shape of the cell 
         include_diag = F)  # remove the diagonal cells 

```
Looks like there are some strong positive correlations (e.g., between Na and Conductance and between N and P) and some strong negative ones (e.g., between N and Density and between P and Density). So what are the actual coefficients (and their significance)?

### (2) A mixed layout



```{r, fig.width = 5, fig.height = 4}

### Mixed layout
# change layout
# tile + p-values + significance stars
# change the labels






```




### (3)

```{r, fig.width = 5, fig.height = 4}

### Modify cell appearance
# highlight cell border


### Cluster the variables and add annotation



```




## Summary




Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

