---
title: "Post #28. Adding an external image to ggplots"
description: |
  Come and learn different ways of adding an external image to ggplots to make your figures more appealing!
author:
  - name: Gen-Chang Hsu
date: 2023-8-22
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
preview: ../../homepage_images/Post28.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Sometimes, we would like to put an image, an illustration, or a silhouette in the figure to make it more visually appealing. How can we do this in ggplots? Well, this is actually not difficult at all. And in this post, I'm going to introduce four ways of adding an external image to ggplots. Read below! 


## Four ways of adding an image to ggplots

### (1) The package `ggimage`

The extension package [`ggimage`](https://yulab-smu.top/pkgdocs/ggimage.html) has a function `geom_image()` that draws images on ggplots. Simply specify the x- and y-coordinates and map the image (either a local file path or an url) to the "image" argument and you're all set!

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)
library(ggimage)

### The url of an iris image
image_url <- "https://images.phylopic.org/images/0a8ab4f9-04c9-4485-b21a-df683d506055/raster/222x512.png"

### A boxplot with an iris image added
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) + 
  geom_boxplot() + 
  geom_image(aes(x = 1, y = 7.5, image = image_url), size = 0.07, color = "purple") +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 13)

```
<br>

We can also use the image as points for the scatterplot:

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_image(aes(image = image_url, color = Species), size = 0.03) + 
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 13)

```

### (2) The package `ggtext`

[`ggtext`](https://wilkelab.org/ggtext/index.html) is another extension package that we can use to add images to ggplots, yet the way how `ggtext` does it is a bit different from `ggimage`: instead of directly specifying the image path in `aes()`, we need to wrap the image path into an HTML \<img\> tag and use the function `geom_richtext(label = )` to render the tag into an actual image for plotting. The size of the image can be controlled using the "width" or "height" attribute in the \<img\> tag . There's one limitation though: you can't change the color of the image as you did earlier. 

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)
library(ggtext)

### Create an HTML <img> tag
image_tag <- "<img src = 'https://images.phylopic.org/images/0a8ab4f9-04c9-4485-b21a-df683d506055/raster/222x512.png' width = '15'/>"

### A boxplot with an iris image added
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) + 
  geom_boxplot() + 
  geom_richtext(aes(x = 1, y = 7.5, label = image_tag), color = NA) +  # specify "color = NA" to remove the outline box 
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 13)

```

As a side note, you can use `ggtext` to modify the labels and text in ggplots in a more sophisticated and customized manner. Check out my [previous post](https://genchanghsu.github.io/ggGallery/posts/2021-07-10-post-5-awesome-text-display-with-ggtext/) on using `ggtext` for awesome text displays in ggplots!

### (3) The package `cowplot`

The third extension package that we can use to add an image to ggplots is [`cowplot`](https://wilkelab.org/cowplot/reference/draw_image.html). The basic idea is that we will first create a drawing canvas using the function `ggdraw()`, and then we can add plots (using the function `draw_plot()`) or images (using the function `draw_image()`) to the canvas. The canvas has x- and y-coordinates between 0 and 1, and we can specify the image position using relative coordinates. (It does require some experimentation to tweak the image position!) Setting "hjust = 0.5" and "vjust = 0.5" will make the center of the image at the point of the specified x- and y-location.

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)
library(cowplot)

### The url of an iris image
image_url <- "https://images.phylopic.org/images/0a8ab4f9-04c9-4485-b21a-df683d506055/raster/222x512.png"

### A boxplot with an iris image added
iris_boxplot <- ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) + 
  geom_boxplot() + 
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 13)

### Draw the plot and image on the canvas
ggdraw() +
  draw_plot(iris_boxplot) +  # the plot will take up the entire canvas space by default
  draw_image(image_url, x = 0.21, y = 0.8, scale = 0.15, hjust = 0.5, vjust = 0.5)  # use the argument "scale" to adjust the size of the image

```

### (4) The package `patchwork`







## Summary

In this post, we looked at four different extension packages that we can use to add an image to ggplots. You may ask: which one should I use? Well, I would say they're pretty much similar in most cases, so simply pick the one that's down your alley! If you want to do something further (e.g., change the image color), then look for a certain package that has the functionality you need. 

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

