---
title: "Post #32. ggplot Map Series No.5: Map annotation with `ggspatial`"
description: |
  Enhance your map with a scale bar and a north arrow using the package `ggspatial`!
author:
  - name: Gen-Chang Hsu
date: 2023-12-24
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Map Series"
preview: ../../homepage_images/Post32.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction




In this post, I'm going to show you how to add them to your map, using the package [`ggspatial`](https://paleolimbot.github.io/ggspatial/index.html). 




## An elevation map of Taiwan



### The boundary map

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)
library(sf)

### Administrative boundary data of Taiwan from the "geoBoundaries" database 
url_Taiwan_boundary_admin0 <- "https://github.com/GenChangHSU/ggGallery/blob/master/_posts/2023-05-24-post-26-ggplot-map-series-no2-simple-feature-maps/geoBoundaries-TWN-ADM0-all.zip?raw=TRUE"
download.file(url_Taiwan_boundary_admin0, "geoBoundaries-TWN-ADM0-all.zip", mode = "wb")
unzip("geoBoundaries-TWN-ADM0-all.zip")

Taiwan_boundary_admin0 <- st_read("./geoBoundaries-TWN-ADM0/geoBoundaries-TWN-ADM0.shp")


```

### Add elevation tiles

```{r, fig.width = 5, fig.height = 4}
library(elevatr)

### Elevation data of Taiwan from the "Amazon Web Services" database  
Taiwan_elevation <- get_elev_raster(locations = Taiwan_boundary_admin0, z = 9, clip = "locations") %>% 
  as.data.frame(xy = T, na.rm = T) %>%
  `colnames<-`(c("lon", "lat", "elev"))


```

### Add a scale bar

```{r, fig.width = 5, fig.height = 4}
library(ggspatial)


  # scalebar(x.min = 117.9, x.max = 118.9, y.min = 21.7, y.max = 22.2,
  #          dist = 50, dist_unit = "km", transform = T, model = "WGS84", 
  #          height = 0.2, st.dist = 0.3, st.size = 2.5) +
  # north(x.min = 118.6, x.max = 118.7, y.min = 22.4, y.max = 22.57, symbol = 10, scale = 5) + 
  # annotate(geom = "text", x = 118.45, y = 22.5, label = "N", size = 5)

```

### Add a north arrow


```{r, fig.width = 5, fig.height = 4}




```




## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!


