---
title: "Post #34. Connecting paired points in ggplots"
description: |
  Check out the handy tips for connecting paired points in ggplots!
author:
  - name: Gen-Chang Hsu
date: 2024-06-11
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2024"
preview: ../../homepage_images/Post34.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Feel so glad (and relieved) to write another blog post after three months of rest! This post was actually inspired by a recent project I am working on: I wanted to make a ggplot comparing the feeding response of control vs. treatment for each larval parent family on two plant types, and I would like to draw a line connecting the control-treatment pair for each family. I thought it was a walk in the park, but when I started making the plot, it turned out to be more complicated than I imagined. Thankfully, after a bit of researching and experimenting, I figured it out and I felt it would be helpful to write something about it. So in this post, I'll show you the tips for connecting paired points in ggplots. Let's do it!

## Case 1. A single group

CO2 dataset
focus on the Q type
X axis is the treatment of one plant species
Y axis is the CO2 uptake rate
Paired by concentration


```{r, fig.width = 5, fig.height = 4}
library(tidyverse)



```




## Case 2. Two or more groups


CO2 dataset
focus on the Q type
X axis is the plant species (three species)
Y axis is the CO2 uptake rate
dodged by treatment
Paired by concentration


```{r, fig.width = 5, fig.height = 4}
library(tidyverse)



```

### Method #1. Use facets







### Method #2. Manually recode the paired points


```{r, fig.width = 5, fig.height = 4}
df <- c("A", "B", "A", "C", "C", "B", "A")
lib <- c("A" = 1, "B" = 2, "C" = 3)
lib[df]



```






## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!





