---
title: "Post #34. Connecting paired points in ggplots"
description: |
  Check out the handy tips for connecting paired points in ggplots!
author:
  - name: Gen-Chang Hsu
date: 2024-06-11
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2024"
preview: ../../homepage_images/Post34.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Feel so glad (and relieved) to write another blog post after three months of rest! This post was actually inspired by a recent project I am working on: I wanted to make a ggplot comparing the feeding response of control vs. treatment for each larval parent family on two plant types, and I would like to draw a line connecting the control-treatment pair for each family. I thought it was a walk in the park, but when I started making the plot, it turned out to be more complicated than I imagined. Thankfully, after a bit of researching and experimenting, I figured it out and I felt it would be helpful to write something about it. So in this post, I'll show you the tips for connecting paired points in ggplots. Let's do it!


## Case 1. A single group

We will use the built-in [CO2](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/CO2) dataset, which contains data from an experiment looking at how chilling affected the CO~2~ uptake rates of the grass species *Echinochloa crus-galli*. 

Let's start with the simplest case where we compare the CO~2~ uptake rates of the chilled vs. nonchilled plant individual #1 at different ambient CO~2~ concentrations. We'll draw a line between each chilled-nonchilled pair to show the treatment effect.

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)

### Select plant individual #1
CO2_plant_1 <- filter(CO2, Plant %in% c("Qn1", "Qc1")) 
  
### Plot  
ggplot(CO2_plant_1) + 
  geom_point(aes(x = Treatment, y = uptake, shape = Treatment, color = factor(conc))) + 
  geom_line(aes(x = Treatment, y = uptake, color = factor(conc), group = factor(conc))) + 
  scale_color_brewer(palette = "Set1", name = "Ambient [CO2]") + 
  theme_classic(base_size = 13)

```

Looks like chilling reduces CO~2~ uptake rates of plant individual #1 under all ambient CO~2~ concentrations.


## Case 2. Two or more groups

Now, what if we're interested in the other plant individuals? Would the patterns be similar? Let's make a plot and see.

```{r, fig.width = 7, fig.height = 4}
### Select plant individual #1 to #3
CO2_plant_1_to_3 <- filter(CO2, Plant %in% c("Qn1", "Qc1", "Qn2", "Qc2", "Qn3", "Qc3")) %>% 
  mutate(Plant_id = str_remove(Plant, "(n|c)")) %>% 
  mutate(Plant_conc_id = str_c(Plant_id, conc, sep = "_"))  # create a unique id for the paired points
  
### Plot  
ggplot(CO2_plant_1_to_3) + 
  geom_point(aes(x = Plant_id, y = uptake, group = Treatment, shape = Treatment, color = factor(conc)),
             position = position_dodge(width = 0.8)) + 
  geom_line(aes(x = Plant_id, y = uptake, group = Plant_conc_id, color = factor(conc)),
            position = position_dodge(width = 0.8)) + 
  scale_color_brewer(palette = "Set1", name = "Ambient [CO2]") + 
  theme_classic(base_size = 13)

```

Oops it doesn't work. How can we solve this problem? Read below!

### Method 1. Use facets

The first solution is to create separate panels for the plant individuals using facets. This is a easy quick fix.

```{r, fig.width = 7, fig.height = 4}
### Plot  
ggplot(CO2_plant_1_to_3) + 
  geom_point(aes(x = Treatment, y = uptake, color = factor(conc))) + 
  geom_line(aes(x = Treatment, y = uptake, group = factor(conc), color = factor(conc))) + 
  facet_wrap( ~ Plant_id) +  # creating facets by plant id
  scale_color_brewer(palette = "Set1", name = "Ambient [CO2]") + 
  theme_classic(base_size = 13)

```

As we can see, all three plant individuals showed similar responses to chilling.

### Method 2. Manually recode the paired points

A second method is to manually recode the x-coordinates of the paired points and join them with lines. This is a bit more tedious, but works if you don't want to use facets. 

Basically, the two treatment groups will be dodged to the two sides for each plant individual, and a unique id is used to connect the paired points. 

```{r, fig.width = 5, fig.height = 4}
df <- c("A", "B", "A", "C", "C", "B", "A")
lib <- c("A" = 1, "B" = 2, "C" = 3)
lib[df]



```






## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!





