---
title: "Post #21. Useful tips for mapping color and fill aesthetics in the same geom layer"
description: |
  A short description of the post.
author:
  - name: Gen-Chang Hsu
date: 2022-10-22
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post21.png
draft: true
---


<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background


## Various mappings of color and fill aesthetics


```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
library(tidyverse)

### fill
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(fill = Species)) +
  scale_fill_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

### color
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species)) +
  scale_fill_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

### both
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(fill = Species, color = after_scale(fill))) +
  scale_fill_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = Species)) +
  scale_fill_brewer(palette = "Set1") + 
  scale_color_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

### lighten color
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(fill = Species, color = after_scale(alpha(fill, 0.1)))) +
  scale_fill_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

### lighten fill
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(fill = Species, fill = after_scale(alpha(fill, 0.4)), color = after_scale(fill))) +
  scale_fill_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

### lighten fill but point too
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = Species), alpha = 0.4) +
  scale_fill_brewer(palette = "Set1") + 
  scale_color_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```


## Summary



Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!







