---
title: "Post #21. Useful tips for mapping color and fill aesthetics in the same geom layer"
description: |
  In this quick post, I will show you how to use color and fill aesthetics together in the same geom layer to achieve the desired outcome!
author:
  - name: Gen-Chang Hsu
date: 2022-10-22
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post21.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 20% !important;
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Background

Color and fill aesthetics often come hand in hand in ggplots, and knowing how to use them together will help achieve the desired plotting outcome and also save lots of your time searching the internet for solutions. In this quick post, I will use a simple boxplot example to demonstrate it.

## Various mappings of color and fill aesthetics

Below is a boxplot showing the sepal length of the three species using the `iris` dataset, with box lines colored by species:

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
library(tidyverse)

ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species)) +
  scale_color_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```

Suppose that we want to fill the boxes by species as well. We can do it by specifying `fill = species` in `aes()` and `scale_fill_brewer(palette = "Set1")`:

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = Species)) +
  scale_color_brewer(palette = "Set1") + 
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```

But there is a shortcut for doing this: the function `after_scale()`. What this function does is that it will apply the scale of a previous aesthetic (in this case "color") to another aesthetic (in this case "fill") so that the latter aesthetic will follow the scale of the previous one. Also, With `after_scale()`, you don't need to specify `scale_fill_brewer(palette = "Set1")` as we did in the previous figure.

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = after_scale(color))) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```


```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = after_scale(color),
                   color = after_scale(alpha(color, 0.1)))) +
  scale_color_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```


```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = after_scale(alpha(color, 0.1)))) +
  scale_color_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```

```{r, fig.width = 6, fig.height = 4, out.width = "75%"}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot(aes(color = Species, fill = after_scale(color)), alpha = 0.1) +
  scale_color_brewer(palette = "Set1") + 
  labs(x = "Species", y = "Sepal length") + 
  theme_classic(base_size = 13)

```


## Summary



Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!







