---
title: "Post #32. ggplot Map Series No.5: Map annotation with `ggsn`"
description: |
  Enhance your map with a scale bar and a north arrow using the package `ggsn`!
author:
  - name: Gen-Chang Hsu
date: 2023-12-24
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
  - "Map Series"
preview: ../../homepage_images/Post32.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Have you ever created a map in ggplot and wondered how to add a scale bar and a north arrow to it? Well, you can easily do it in R using the package [`ggsn`](https://paleolimbot.github.io/ggspatial/index.html), and this is the topic for this short post.


## An annotated elevation map of Taiwan

### (1) Create an elevation map
We'll create an elevation map of Taiwan as our example. Here, we'll use the function `get_elev_raster()` from the package `elevatr` to retrieve the elevation tiles of Taiwan from the online map databases. Basically, what we need to do is to read in a shapefile of the area of interest and pass it to the function. It will return a "RasterLayer" object, which is then converted to a dataframe for plotting in ggplot.

```{r, fig.width = 3.5, fig.height = 3.5, out.width = "60%"}
library(tidyverse)

### Download and read in the shapefile of Taiwan's administrative boundary
### The zipped shapefile is available in the folder of this post on GitHub
library(sf)

Taiwan_boundary_admin0_url <- "https://github.com/GenChangHSU/ggGallery/blob/master/_posts/2023-12-24-post-32-ggplot-map-series-no5-map-annotation-with-ggspatial/geoBoundaries-TWN-ADM0-all.zip?raw=TRUE"
download.file(Taiwan_boundary_admin0_url, "geoBoundaries-TWN-ADM0-all.zip", mode = "wb")
unzip("geoBoundaries-TWN-ADM0-all.zip", exdir = "geoBoundaries-TWN-ADM0-all")

Taiwan_boundary_admin0 <- st_read("geoBoundaries-TWN-ADM0-all/geoBoundaries-TWN-ADM0.shp", quiet = T)

### Get the elevation data of Taiwan
library(elevatr)
library(terra)  # for the function as.data.frame()

Taiwan_elevation <- get_elev_raster(locations = Taiwan_boundary_admin0, z = 9, clip = "locations") %>%  # z controls the resolution of the elevation tiles
  as.data.frame(xy = T, na.rm = T) %>%  # convert the RasterLayer to a dataframe
  `colnames<-`(c("lon", "lat", "elevation")) %>% 
  mutate(elevation = if_else(elevation < 0 | is.na(elevation), 0, elevation))  # set the negative elevation values and NAs to 0

### Make the map
library(tidyterra)  # for the function hypso.colors()
topo_pal <- hypso.colors(n = 10, palette = "dem_poster")  # a topographic color palette

Taiwan_elevation_map <- ggplot() + 
  geom_raster(data = Taiwan_elevation, aes(x = lon, y = lat, fill = elevation), show.legend = F) +
  scale_fill_gradientn(colors = topo_pal,
                       values = c(seq(0, 0.35, length.out = 5), seq(0.5, 1, length.out = 5))) +
  coord_sf(xlim = c(119, 123), ylim = c(21.5, 25.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank())

Taiwan_elevation_map

```

### (2) Add a scale bar

It's time to add a scale bar! `ggsn` has the function `scalebar()` to add and customize a scale bar:

```{r, fig.width = 3.5, fig.height = 3.5, out.width = "60%"}
library(ggsn)

Taiwan_elevation_map_annotated <- Taiwan_elevation_map + 
  scalebar(x.min = 121.5, x.max = 122.5, y.min = 21.7, y.max = 22.2,
           dist = 50, dist_unit = "km", transform = T, model = "WGS84",
           height = 0.2, st.dist = 0.3, st.size = 2.5)

Taiwan_elevation_map_annotated

```


### (3) Add a north arrow


```{r, fig.width = 3.5, fig.height = 3.5, out.width = "60%"}


  # north(x.min = 118.6, x.max = 118.7, y.min = 22.4, y.max = 22.57, symbol = 10, scale = 5) + 
  # annotate(geom = "text", x = 118.45, y = 22.5, label = "N", size = 5)

```


### (4) Polish the map


```{r, fig.width = 3.5, fig.height = 3.5, out.width = "60%"}




```



## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!


