---
title: "Post #29. Bad weather today—Creating raincloud plots in ggplots"
description: |
  Come and learn how to create raincloud plots in ggplots to visualize data distribution.
author:
  - name: Gen-Chang Hsu
date: 2023-09-09
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2023"
preview: ../../homepage_images/Post29.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
    color: #2780e3 !important;
      border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 25% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Rainclound plot, what a cool name! You might, or might not have heard of it before, but as you'll see soon, this type of chart is awesome for visualizing data distribution.

A rainclound plot is actually not a single plot, but instead it consists of three different plots: a boxplot, a dot plot, and a density plot. With these components, a rainclound plot can capture various aspects of the data and convey more information than the individual constituent plots alone.

This post is actually inspired by previous two posts ([post 1](https://3mw.albert-rapp.de/p/rainclouds-ggplot?utm_source=3mw.albert-rapp.de&utm_medium=newsletter&utm_campaign=3mw-rainclouds-with-ggplot-a-great-way-to-visualize-distributions) and [post 2](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-and-how-to-create-them-with-ggplot2/?utm_source=3mw.albert-rapp.de&utm_medium=newsletter&utm_campaign=3mw-rainclouds-with-ggplot-a-great-way-to-visualize-distributions)) I stumbled upon the other day. Both posts give a neat tutorial of creating rainclound plots using the extension package `ggdist`, so I will not repeat that again here. Instead, what I'm going to do is to show you how to create a raincloud plot in ggplots using basic ggplot functions. Should be fun!

## Create the plot

We'll use the `penguins` dataset in the package `palmerpenguins`, which contains 

### The box plot

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)
library(palmerpenguins)

P_raincloud <- penguins %>% 
  mutate(species = fct_relevel(species, c("Chinstrap", "Adelie", "Gentoo"))) %>% 
  filter(!is.na(sex)) %>%  
  ggplot() + 
  geom_boxplot(aes(x = species, fill = species, y = body_mass_g), 
               width = 0.15,
               show.legend = F)

P_raincloud

```

### The dot plot

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P_raincloud <- P_raincloud +
  geom_dotplot(aes(x = species, fill = species, y = body_mass_g),
               binaxis = "y",
               position = position_nudge(x = 0.15),
               dotsize = 0.5,
               show.legend = F)

P_raincloud

```

### The density plot

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
P_raincloud <- P_raincloud + 
  geom_ribbon(aes(x = species, y = body_mass_g, 
                  xmin = after_stat(group),
                  xmax = after_stat(group - ndensity*0.2),
                  colour = species, fill = after_scale(alpha(colour, 0.3))),
              stat = "density", outline.type = "upper",
              position = position_nudge(x = -0.1),
              show.legend = F) +
  coord_cartesian(x = c(1, 3)) + 
  theme_classic(base_size = 13)

P_raincloud 

```


## Summary




Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!

