---
title: "Post #10. Tweak the legends in ggplots with guides()"
description: |
  In this post, you will learn how to tweak the appearance of the ggplot legends using with the function `guides()`.
author:
  - name: Gen-Chang Hsu
date: 2022-01-22
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2022"
preview: ../../homepage_images/Post10.png
draft: true
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

<!-- Start of the article -->

Welcome to the first post in 2022—also my tenth ggGallery post! A small milestone reached. Keep going and head towards the next one!

## Introduction



## Discrete-scale legends
We can modify the legends mapped to a continuous variable using `guide_legend()`

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}
library(tidyverse)

```


## Continuous-scale legends
We can modify the legends mapped to a continuous variable using `guide_colorbar()`

```{r, fig.width = 5, fig.height = 3, out.width = "70%"}


```


## Summary

In this post, we've seen

Another way to modify the legend appearance is using the `theme()`. 





Hope you enjoy the reading and don't forget to leave your comments and suggestions below if you have any!
