---
title: "Post #40. Mathematical expressions in ggplots"
description: |
  A short description of the post.
author:
  - name: Gen-Chang Hsu
date: 2025-03-28
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2025"
preview: ../../homepage_images/Post40.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Mathematical expressions are super common in figures: adding stats results or text labels with subscripts and superscripts, using Greek letters and math symbols in axis titles, just to name a few. In this short post, I'm going to show you several ways to annotate math expressions in ggplots. Read on!


## Mathematical expressions in ggplots 

### (1) Math expressions using expression()

We can use the function `expression()` to create a wide variety of math expressions: italic and bold fonts, superscripts and subscripts, Greek letters, hat and bar symbols, fractions, integrals and derivatives, etc. Here are some commonly used ones (see `?plotmath` for a complete list of math expressions!): 

```{r}
library(tidyverse)

### Various math expressions and symbols
ggplot() + 
  annotate(geom = "text", x = 1, y = 1, label = expression(Italic~and~bold~font:~This~is~italic(italic~R)~and~bold(bold~R)~and~bolditalic(bold~italic~R)), size = 5, hjust = 0) + 
  annotate(geom = "text", x = 1, y = 0.8, label = expression(Superscript~and~subscript:~italic(F[1*","*20]) == 5.2*";"~italic(R^2) == 0.8), size = 5, hjust = 0) + 
  annotate(geom = "text", x = 1, y = 0.6, label = expression(Greek~letter:~italic(alpha) == 0.05*";"~italic(chi^2) == 3.5), size = 5, hjust = 0) + 
  annotate(geom = "text", x = 1, y = 0.4, label = expression(Math~Annotation:~y == frac(1, sigma*sqrt(2*pi))~e^-frac((x-mu)^2, 2*sigma^2)), size = 5, hjust = 0) + 
  annotate(geom = "text", x = 1, y = 0.2, label = expression(Calculus:~italic(frac(partialdiff*N, partialdiff*t))*";"~italic(integral(f(x)*dx, a, b))), size = 5, hjust = 0) + 
  scale_x_continuous(limits = c(0.9, 10)) +
  scale_y_continuous(limits = c(0, 1.1)) + 
  theme_void() +
  theme(plot.background = element_rect())

```


We can use objects in math expressions too. However, `expression()` can't handle that. Instead, we use the sister function `bquote()`: the syntax is the same as that for `expression()`, but now you can enclose the objects in ".()" and the object values will show up in the text!

As an example, we can extract the linear model results, store them as objects, and add text annotations to the plot using `bquote()`: 

```{r}
### Create functions that extract linear model results
# function for extracting coefficients
lm_coefficients_fun <- function(model){
  
  a <- round(coef(model)[1], 2)
  b <- round(coef(model)[2], 2)
  
  if (b > 0) {
    
    # use ".()" to substitute object values in the expression
    equation <- bquote(italic(y) == .(a) + .(b)*x)
    
  } else {
    
    b <- abs(b)
    equation <- bquote(italic(y) == .(a) - .(b)*x)
    
  }
  
  return(equation)
}

# function for extracting statistical results
lm_stat_results_fun <- function(model){
  
  f <- round(summary(model)$fstatistic[1], digits = 2)
  df1 <- round(summary(model)$fstatistic[2], 1)
  df2 <- round(summary(model)$fstatistic[3], 1)
  p_value <- round(anova(model)$`Pr(>F)`[1], 3)
  r2 <- round(summary(model)$r.squared, digits = 2)
  
  
  if (p_value > 0.001) {
    
    # use ".()" to substitute object values in the expression
    statistics <- bquote(italic(F[.(df1)*","*.(df2)]) == .(f)*","~italic(p) == .(p_value)*","~italic(R^2) ==.(r2))
    
  } else {
    
    statistics <- bquote(italic(F[.(df1)*","*.(df2)]) == .(f)*","~italic(p) < 0.001*","~italic(R^2) == .(r2))
    
  }
  
  return(statistics)
}

### Extract model results and store the expressions as objects
lm_iris <- lm(Sepal.Width ~ Sepal.Length, data = iris)
coefs <- lm_coefficients_fun(model = lm_iris)
stats <- lm_stat_results_fun(model = lm_iris)

### Use "bquote()" again to add the expressions to the plot
ggplot() +
  annotate(geom = "text", x = 1, y = 1, label = bquote(.(coefs)), size = 5, hjust = 0) + 
  annotate(geom = "text", x = 1, y = 0.9, label = bquote(.(stats)), size = 5, hjust = 0) + 
  scale_x_continuous(limits = c(0.9, 10)) +
  scale_y_continuous(limits = c(0.6, 1.1)) + 
  theme_void() +
  theme(plot.background = element_rect())

```


### (2) Math expressions using the package `ggtext`

If you have a keen eye, you might have noticed that the words in *italic* and _**bold italic**_ in the first example look a bit funky: the letters seem to be spaced out a bit. 

This is totally fine, but we can make the words look better by use the package `ggtext` (check out my [previous post](https://genchanghsu.github.io/ggGallery/posts/2021-07-10-post-5-awesome-text-display-with-ggtext/) on this!) using HTML syntax.

Another benefit of using `ggtext` is that you can insert line breaks between text, which `expression()` can't.

```{r}
library(ggtext)

ggplot() +
  geom_richtext(aes(x = 1, y = 1), label = "This is *italic R* and **bold R** and _**bold italic R**_", size = 5, hjust = 0, fill = NA, label.color = NA) + 
  geom_richtext(aes(x = 1, y = 0.8), label = "*F*<sub><span style='font-size: 8pt'>1,20</span></sub> = 5.2; *R<sup><span style='font-size: 10pt'>2</span></sup>* = 0.8", size = 5, hjust = 0, fill = NA, label.color = NA) + 
  geom_richtext(aes(x = 1, y = 0.6), label = "*&alpha;* = 0.05; *&chi;<sup><span style='font-size: 10pt'>2</span></sup>* = 3.5", size = 5, hjust = 0, fill = NA, label.color = NA) + 
  geom_richtext(aes(x = 1, y = 0.4), label = "This is a line break: <br> *p* = 0.001", size = 5, hjust = 0, fill = NA, label.color = NA) + 
  scale_x_continuous(limits = c(0.9, 10)) +
  scale_y_continuous(limits = c(0, 1.1)) + 
  theme_void() +
  theme(plot.background = element_rect())

```

However, `ggtext` also has its limitations: currently it only supports a limited number of HTML <tags>, and it is a bit more difficult to work with math symbols like fraction, integral, and partial derivative.


## Summary

To recap, we learned how to add math expressions to ggplots using `expression()`, `bquote()`, and the package `ggtext`. Knowing the strength of each method and when to use it will give you the best outcome!

Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!





