---
title: "Post #5. Awesome text displays with ggtext"
description: |
  Enhance the text displays in your ggplots with the extension package ggtext!
author:
  - name: Gen-Chang Hsu
date: 2021-07-10
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2021"
preview: ../../homepage_images/Post5.png  
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

<!-- Start of the article -->

## Overview
Text is one of the fundamental elements in a graph. A well-designed text display can make your plot look pleasant and more informative. ggplot2 provides many functions and arguments for text manipulation (appearance, layout, annotation, etc.), which are already more than enough most of the time. But if you want to do something further, the extension package [`ggtext`](https://wilkelab.org/ggtext/index.html) has more to offer. This package has several added functionalities currently not implemented in ggplot2, the most important being that it supports the use of Markdown/HTML/CSS, which allows for greater flexibility in text rendering. So in this post, I will be showing you some of these cool features. Ready? Let's get started!  


## Date preparation
We will be using the data from the [International Biology Olympiad](https://www.ibo-info.org/en/contest/past-ibos.html) website. IBO is arguably the most magnificent annual biology event for high school students around the world. Since the very first one held 30 years ago in the Czech Republic, more countries have joined as members, and it would be interesting to see how the numbers of participants changed over the past three decades.

To do so, we will first scrap the data from the website using the package [`rvest`](https://rvest.tidyverse.org/), and then do some data manipulation for later plotting use. Since web-scraping is not the main focus of this post, I will not go into the details. If interested, you can visit the package's website for more information. There are also plenty of learning resources out there on the internet. 

```{r, fig.width = 5, fig.height = 3, out.width = "70%", fig.show = "hold"}
library(tidyverse)
library(rvest)

# The website url
IBO_url <- "https://www.ibo-info.org/en/contest/past-ibos.html"

# Read the html and select elements using css selectors
IBO_html <- read_html(IBO_url) %>%
  html_elements("div.item__content") %>%
  html_text2()

# Extract the information from the html text
Year <- IBO_html %>% str_extract("[:digit:]{4}(?=:)")

Country <- IBO_html %>% str_extract("(?<=,\\s)[:alpha:]+[:blank:]*[:alpha:]*") %>%
  replace_na("Japan")

Students <- IBO_html %>% str_extract("[:digit:]*(?=\\s*students)")

# Put all information in a dataframe  
IBO_data <- tibble(Year, Country, Students) %>%
  mutate(Year = as.numeric(Year),
         Students = as.numeric(Students))

```
<br>

## The basic plot
Now we have our dataset at hand. It's time to visualize it using a simple line chart:

```{r, fig.width = 5, fig.height = 3, out.width = "85%", fig.show = "hold"}
ggplot(IBO_data, aes(x = Year, y = Students)) + 
  geom_point() + 
  geom_line() + 
  labs(y = "Number of participants") +
  theme_classic() 

```

As you can see, the numbers of participants have generally increased over time, but there was a sharp drop in 2020 due to the COVID-19 global pandemic, which might have caused some countries to withdraw from the event.


## Make it texty 
The above line chart looks fine, though a bit plain. So now we will make use of `ggtext` to add some flavor to it. Also, I will split the line chart into two panels (one from 1990 to 2005 and the other from 2006 to 2020) using facets for two reasons: (1) to avoid crowding and overlapping, and (2) to demonstrate how to modify the facet strip text. 

There are three main parts we can do with `ggtext`:<br>

<span style="display: block; margin-top: 0px; margin-left: 10px"> **(1) Modify text outside the plot area**</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> **(2) Modify text inside the plot area**</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> **(3) Add external images to the plot**</span>

I will go through the details in the following sections.


### 1. Modify text outside the plot area
First, we will begin by modifying the text outside the plot area — the title, axis labels, captions, facet strips, and so on.   

```{r, fig.width = 6, fig.height = 6, out.width = "85%", fig.show = "hold"}
library(ggtext)
library(extrafont) # Fonts for plotting
# font_import() # Import system fonts (only need to run it once on your device) 
loadfonts(quiet = T) # Register system fonts (need to run it in every R session)

# Create a new column for the two time periods
IBO_data <- IBO_data %>% 
  mutate(Time_period = ifelse(Year <= 2005, "1990 to 2005", "2006 to 2020"))

# Plain faceted line chart
IBO_plot <- ggplot(IBO_data, aes(x = Year, y = Students)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Time_period, scales = "free", nrow = 2, strip.position = "right") +
  scale_x_continuous(breaks = 1990:2020) +
  scale_y_continuous(expand = c(0, 50)) + 
  theme_classic()

# Title
IBO_plot <- IBO_plot + 
  labs(title = "<span style = 'font-size: 18pt'><i>**I**nternational **B**iology **O**lympiad<i/></span><br>Numbers of participants over past three decades") + 
  theme(plot.title = element_textbox_simple(size = 10,
                                            family = "Century Gothic",
                                            color = "#007575",
                                            fill = "#e0f2f2",
                                            halign = 0.5,
                                            lineheight = 1.5,
                                            padding = margin(5, 1, 5, 1), 
                                            margin = margin(0, 0, 5, 0)))

# Axis labels
IBO_plot <- IBO_plot + 
  labs(x = "", y = "Number of participants<br><span style = 'font-size: 9pt;'>(**Note:** different y-axis ranges in the two panels)</span>") + 
  theme(axis.title.y = element_textbox_simple(size = 14,
                                              family = "Arial",
                                              color = "#045a8d",
                                              fill = "#fff080",
                                              halign = 0.5,
                                              orientation = "left-rotated",
                                              r = unit(5, "pt"),
                                              padding = margin(4, 4, 4, 4),
                                              margin = margin(0, 0, 8, 0),
                                              minwidth = unit(2, "in"),
                                              maxwidth = unit(5, "in")))

# Caption
IBO_plot <- IBO_plot + 
  labs(caption = "<i><span style = 'font-size: 9pt;'>Data source: https:&#47;&#47;www&#46;ibo-info.org/en/contest/past-ibos.html</span><br>Retrieved July, 2021<i/>") + 
  theme(plot.caption = element_markdown(size = 8,
                                        family = "serif",
                                        color = "grey50",
                                        lineheight = 1.5))

# Facet strips
IBO_plot <- IBO_plot + 
  theme(strip.background = element_blank(),
        strip.text.y = element_textbox_simple(size = 12,
                                              family = "Comic Sans MS",
                                              color = "white", 
                                              fill = "#810f7c", 
                                              halign = 0.5,
                                              orientation = "right-rotated",
                                              r = unit(8, "pt"),
                                              padding = margin(1, 1, 1, 1),
                                              margin = margin(3, 3, 3, 3)))
                                     
IBO_plot

```

The key functions here are `element_textbox_simple()` and `element_markdown()`. Basically they do almost the same job, the main difference being that the former allows users to have more control over the appearance of the text box, whereas the latter focuses on modifying the text itself. So if you are working with text only, you can simply use `element_markdown()`.

The usage of `element_textbox_simple()` and `element_markdown()` are pretty much similar to their ggplot2 counterpart `element_text()`. When you want to modify a specific part of the plot (e.g., title), just call these functions to the corresponding arguments (e.g., `plot.title = `) in the `theme()`.  

Also, you might have seen that I use quite a bit of the Markdown/HTML/CSS syntaxes. Remember in the beginning, I mentioned that `ggtext` can render text written in Markdown/HTML/CSS languages. With these, you can easily do some advanced text manipulation, for example, modifying the appearance of a specific letter/word in a sentence.


### 2. Modify text inside the plot area
After getting the text outside the plot done, we can next work on the text inside the plot area. Specifically, I will add some labels besides points and modify their appearance a bit.

```{r, fig.width = 6, fig.height = 6, out.width = "85%", fig.show = "hold"}
# Add country labels to the points
IBO_plot <- IBO_plot +
  geom_richtext(aes(label = Country), 
                size = 2,
                color = colorRampPalette(c("#a50f15", "#e66101"))(nrow( IBO_data)),
                fill = "transparent",
                angle = 45,
                hjust = 0.1,
                vjust = -0.1,
                label.r = unit(4, "pt"),
                label.padding = unit(c(1, 5, 1, 5), "pt"))

IBO_plot

```

Here, I use the function `geom_richtext()` to map the labels to the points. This function is quite similar to ggplot2's `geom_text()` or `geom_label()`, but it has an additional feature: the text/box can be rotated by any angles.


### 3. Add external images to the plot
Now we have already done with our text manipulation. Most of the time, we will end it here and happy to go. However, there is one extra thing we can do to make the plot fancier — adding external images to the plot! Of course you can do this manually later (perhaps in PowerPoint), but isn't it great to complete all the work at once in R? This will streamline the process and make it fully reproducible! So let's do it!

The principle of adding images is the use of HTML `<img>` tag, which has a basic syntax structure `<img scr = 'path_to_image'/>`. If you are not familiar with HTML, don't panic. Simply copy and paste the code in the example below and replace the image path with yours (can be a website url or a local directory). The additional `width` attribute in the `<img>` tag allows for controlling the width of the image.

```{r, fig.width = 6, fig.height = 6, out.width = "85%", fig.show = "hold"}
# Download the zip file of the logos from my GitHub repository
# The file will be saved in the current directory
download.file(url = "https://raw.githubusercontent.com/GenChangHSU/ggGallery/master/_posts/2021-07-10-post-5-awesome-text-display-with-ggtext/IBO_logos.zip", destfile = "./IBO_logos.zip")

# Unzip the file
unzip(zipfile = "./IBO_logos.zip")

# HTML <img> tags
IBO_logo <- rev(paste0("<img src = './IBO_logos/", Year, ".png' width = '15' /><br>", Year))

# Add the logos to the x-axis 
IBO_plot <- IBO_plot +
  scale_x_continuous(breaks = 1990:2020, labels = IBO_logo) +
  theme(axis.text.x = element_markdown(color = "black", size = 7))

# Add another logo to the upper panel
IBO_logo_df <- data.frame(x = 1992, 
                          y = 200,
                          Time_period = "1990 to 2005",
                          logo = paste0("<img src = './IBO_logos/main.png' width = '100' />"))

IBO_plot <- IBO_plot +
  geom_richtext(data = IBO_logo_df, 
                aes(x = x, y = y, label = logo), 
                colour = "transparent")

IBO_plot

```

As shown above, you can add external images to two areas of a plot: the axes and the plot panel. 

For the first one, you need to specify in `scale_xx_XXX()` the positions at which the images should be placed on the axes (`breaks = `), along with the HTML `<img>` tags as labels (`labels = `). After that, call `element_markdown()` to the corresponding theme element (in this example `axis.text.x`) in the `theme()`. The images will then be displayed along the axes. 

For the second one, simply create a dataframe containing the x and y positions as well as a column containing the HTML `<img>` tags, and use `geom_richtext()` to map the image tags to the plot. The images will then be displayed at the designated positions.


## Summary
A quick recap of what we have done throughout the post. **First**, we modified the plot title, axis labels, caption, and facet strips using two `theme()` elements `element_textbox_simple()` and `element_markdown()`. **Second**, we added some labels to the plot using the geom layer `geom_richtext()` and customized the labels a bit. **Third**, we used `element_markdown()` to add external images to the x-axis, and `geom_richtext()` to add another image to the plot area.

There are lots of things you can do for your text using `ggtext`. Spend some time exploring the package website. I believe you will become a master of text. By the way, I highly recommend having a basic understanding of Markdown/HTML/CSS, which will allow you to fully leverage the strength of `ggtext`!

Oh, one last thing forget to say: I was one of the participants in the 2014 IBO in Bali, Indonesia. It's been seven years. Time flies! In fact, this is one of the main reasons why I chose IBO data as the example dataset. The other is that the website contains many logos, which serve well the purpose of this post.

That's it for this post and don't forget to leave your comments and suggestions below if you have any!


