---
title: "Post #5. Awesome text displays with ggtext"
description: |
  Enhance the text displays in your ggplots with the extension package `ggtext`!
author:
  - name: Gen-Chang Hsu
date: 2021-07-10
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2021"
preview: ../../homepage_images/Post5.png  
---

<!-- Webpage style -->
<style>
d-article a {
    color: #2780e3 !important;
    border-bottom: none !important;
}

d-article a:hover {
    color: #2780e3 !important;
    border-bottom: 2px solid !important;
}

d-byline {
    margin-left: -10% !important;
}

d-title {
    margin-left: -10%;
}

d-title h1{
    font-size: 45px;
    width: 130%;
}

d-article {
    margin-left: -10%;
}

p {
    width: 120%;
}

d-article h2 {
    width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
    width: 120%;
}

d-article div.sourceCode { 
    width: 120% !important;
}

d-article pre { 
    width: 120% !important;
}

div.l-body {
    width: 120% !important;
}

</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

<!-- Start of the article -->

## Overview
Text is one of the fundamental elements in a graph. A well-designed text display can make your plot look pleasant and more informative. ggplot2 provides many functions and arguments for text manipulation (appearance, layout, annotations, etc.), which are already more than enough most of the time. But if you want to do something further, the extension package [`ggtext`](https://wilkelab.org/ggtext/index.html) has more to offer. This package has several added functionalities currently not implemented in ggplot2, the most important being that it supports the use of Markdown/HTML/CSS, which allows for greater flexibility in text rendering. So in this post, I will be showing you some of these cool features. Ready? Let's get started!  


## Date preparation
We will be using the data from the [International Biology Olympiad](https://www.ibo-info.org/en/contest/past-ibos.html) website. IBO is arguably the most magnificent annual biology event for high school students around the world. Since the very first one held 30 years ago in the Czech Republic, more countries have joined as members, and it would be interesting to see how the numbers of participants changed over the past three decades.

To do so, we will first scrap the data from the website using the package [`rvest`](https://rvest.tidyverse.org/), and then do some data manipulation for later plotting use. Since web-scraping is not the main focus of this post, I will not go into the details. If interested, you can visit the package's website for more information. There are also plenty of learning resources out there on the internet. 

```{r, fig.width = 5, fig.height = 3, out.width = "70%", fig.show = "hold"}
library(tidyverse)
library(rvest)

# The website url
IBO_url <- "https://www.ibo-info.org/en/contest/past-ibos.html"

# Read the html and select elements using css selectors
IBO_html <- read_html(IBO_url) %>%
  html_elements("div.item__content") %>%
  html_text2()

# Extract the information from the html text
Year <- IBO_html %>% str_extract("[:digit:]{4}(?=:)")

Country <- IBO_html %>% str_extract("(?<=,\\s)[:alpha:]+[:blank:]*[:alpha:]*") %>%
  replace_na("Japan")

Students <- IBO_html %>% str_extract("[:digit:]*(?=\\s*students)")

# Put all information in a dataframe  
IBO_data <- tibble(Year, Country, Students) %>%
  mutate(Year = as.numeric(Year),
         Students = as.numeric(Students))

```

## The basic plot
Now we have our dataset at hand. It's time to visualize it using a simple line chart:

```{r, fig.width = 5, fig.height = 3, out.width = "85%", fig.show = "hold"}
ggplot(IBO_data, aes(x = Year, y = Students)) + 
  geom_point() + 
  geom_line() + 
  labs(y = "Number of participants") +
  theme_classic() 

```

As you can see, the numbers of participants have generally increased over time, but there was a sharp drop in 2020 due to the COVID-19 global pandemic, which might have caused some countries to withdraw from the event.


## Make it texty 
The above line chart looks fine, though a bit plain. So now we will make use of `ggtext` to add some flavor to it. Also, I will split the line chart into two panels (one from 1990 to 2005 and the other from 2006 to 2020) using facets for two reasons: (1) to avoid crowding and overlapping, and (2) to demonstrate how to modify the facet strip text. 

There are three main parts we can do with `ggtext`:<br>

<span style="display: block; margin-top: 0px; margin-left: 10px"> **(1) Modify text outside the plot area**</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> **(2) Modify text inside the plot area**</span>
<span style="display: block; margin-top: 5px; margin-left: 10px"> **(3) Add images to the plot**</span>

I will go through the details in the following sections.


### 1. Modify text outside the plot area
First, we can modify the text outside the plot area — the title, axis labels, captions, facet strip text, and so on.   

```{r, fig.width = 6, fig.height = 6, out.width = "85%", fig.show = "hold"}
library(ggtext)
library(extrafont) # Fonts for plotting
font_import() # Import the fonts
loadfonts(device = "win") # Load in the fonts

# Split into two time periods
IBO_data <- IBO_data %>% 
  mutate(Time_period = ifelse(Year <= 2005, "1990 to 2005", "2006 to 2020"))

# Faceted line chart
IBO_plot <- ggplot(IBO_data, aes(x = Year, y = Students)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Time_period, scales = "free", nrow = 2, strip.position = "right") +
  scale_x_continuous(breaks = 1990:2020) +
  scale_y_continuous(expand = c(0, 50)) + 
  theme_classic()

# Title
IBO_plot <- IBO_plot + 
  labs(title = "<span style = 'font-size: 18pt'><i>**I**nternational **B**iology **O**lympiad<i/></span><br>Numbers of participants over past three decades") + 
  theme(plot.title = element_textbox_simple(size = 10,
                                            halign = 0.5,
                                            color = "#007575",
                                            family = "Century Gothic",
                                            lineheight = 1.5,
                                            padding = margin(5, 1, 5, 1), 
                                            margin = margin(0, 0, 5, 0),
                                            fill = "#e0f2f2"))

# Axis labels
IBO_plot <- IBO_plot + 
  labs(x = "", y = "Number of participants<br><span style = 'font-size: 9pt;'>(**Note:** different y-axis ranges in the two panels)</span>") + 
  theme(axis.title.y = element_textbox_simple(size = 14, 
                                              halign = 0.5,
                                              color = "#045a8d",
                                              family = "Arial",
                                              orientation = "left-rotated",
                                              padding = margin(4, 4, 4, 4),
                                              margin = margin(0, 0, 8, 0),
                                              minwidth = unit(2, "in"),
                                              maxwidth = unit(5, "in"),
                                              r = unit(5, "pt"),
                                              fill = "#fff080"))

# Caption
IBO_plot <- IBO_plot + 
  labs(caption = "<i><span style = 'font-size: 9pt;'>Data source: https:&#47;&#47;www&#46;ibo-info.org/en/contest/past-ibos.html</span><br>Retrieved July, 2021<i/>") + 
  theme(plot.caption = element_markdown(size = 8, 
                                        lineheight = 1.5,
                                        family = "serif",
                                        color = "grey50"))

# Facet strip text
IBO_plot <- IBO_plot + 
  theme(strip.background = element_blank(),
        strip.text.y = element_textbox_simple(size = 12,
                                              color = "white", 
                                              fill = "#810f7c", 
                                              family = "Comic Sans MS",
                                              orientation = "right-rotated",
                                              halign = 0.5, 
                                              r = unit(8, "pt"), 
                                              width = unit(1, "npc"),
                                              padding = margin(1, 1, 1, 1),
                                              margin = margin(3, 3, 3, 3)))
                                     
IBO_plot

```

So you can see that I use quite a bit of these syntaxes. Remember in the beginning, I mentioned that `ggtext` can render text written in Markdown/HTML/CSS language. With these, you can easily modify the appearance of a specific letter/word in a sentence.


### 2. Modify text inside the plot area
Next, we will shift gears a bit and focus on the text inside the plot. Specifically, I will add labels besides the points.

```{r, fig.width = 6, fig.height = 6, out.width = "85%", fig.show = "hold"}
IBO_plot <- IBO_plot +
  geom_richtext(aes(label = Country), 
                angle = 45, 
                size = 2,
                hjust = 0.1,
                vjust = -0.1,
                label.padding = unit(c(1, 5, 1, 5), "pt"),
                label.r = unit(4, "pt"),
                fill = "transparent",
                color = colorRampPalette(c("#a50f15", "#e66101"))(nrow( IBO_data)))

IBO_plot

```


### 3. Add images to the plot
Finally, we can add images to the plot using the HTML `<img>` tag:

```{r, fig.width = 6, fig.height = 6, out.width = "85%", fig.show = "hold"}
# The url for the IBO logos (available on my GitHub repository)
IBO_logo <- rev(paste0("<img src = './IBO_logos/", Year, ".png' width = '15' /><br>", Year))

# Add the IBO logos to the x-axis 
IBO_plot <- IBO_plot +
  scale_x_continuous(breaks = 1990:2020, labels = IBO_logo) +
  theme(axis.text.x = element_markdown(color = "black", size = 7))

# Add the main logo to the upper panel
IBO_plot + 
  geom_richtext(data = data.frame(x = c(1992, 2007), 
                                  y = c(200, 300),
                                  Time_period = c("1990 to 2005", "2006 to 2020"),
                                  logo = c("<img src = './IBO_logos/main.png' width = '100' />", "")),
                aes(x = x, y = y, label = logo), 
                colour = "transparent")


```




## Summary
There are lots of things you can do for your text using `ggtext`. Spend some time exploring the package website. I believe you will become a master of text. By the way, I highly recommend having a basic understanding of Markdown/HTML/CSS, which will allow you to fully leverage the strength of `ggtext`!

Oh, one last thing forget to say: I was one of the participants in the 2014 IBO in Bali, Indonesia It's been seven years. Time flies! In fact, this is one of the main reasons why I chose IBO data as the example dataset. The other is that the website contains many logos, which serve well the purpose of this post.

That's it for this post and don't forget to leave your comments and suggestions below if you have any!


