---
title: 'Post #33. A "timely" plot—Visualizing time series data in ggplots'
description: |
  Create awesome time series line charts in ggplots.
author:
  - name: Gen-Chang Hsu
date: 2024-02-21
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2024"
preview: ../../homepage_images/Post33.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
  color: #2780e3 !important;
  border-bottom: none !important;
}

d-article a:hover {
  color: #2780e3 !important;
  border-bottom: 2px solid !important;
}

d-byline {
  margin-left: -10% !important;
}

d-title {
  margin-left: -10%;
}

d-title h1{
  font-size: 45px;
  width: 130%;
}

d-article {
  margin-left: -10%;
}

p {
  width: 120%;
}

d-article h2 {
  width: 120%;
}

d-article h3 {
  width: 120%;
  font-size: 28px;
}

div.article-footer{
  width: 120%;
}

d-article div.sourceCode { 
  width: 120% !important;
}

d-article pre { 
  width: 120% !important;
}

div.l-body {
  width: 120% !important;
}

d-article table thead tr {
  height: 60px !important;
  width: 210px !important;
}

d-article table thead th {
  height: 60px !important;
  font-size: 22px !important;
  width: 210px !important;
}

d-article table tbody tr td {
  font-size: 21px !important;
  height: 60px !important;
  text-align: center !important;
  width: 210px !important;
}

div.l-body img {
    margin-right: 30% !important;
}

</style>

<head>
  <base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

Welcome to the first post of 2024! It's been a while since my last post last year (to be precise two months; maybe it's not really long?), and I feel excited to be back in the saddle and write something about ggplots. Hope we'll have another year of posts full of useful and fun topics! 

To kick things off, I'll be showing you how to create time series plots in ggplots. Time series plots are essentially a type of line charts with "time" on the x-axis. Of course, you can simply treat time as a numeric variable and plot it as a usual line chart, but ggplot does offer a set of functions that work directly with time. So in this post, we're going to explore these functions and see how we can leverage them to create a nice time series plot! 


## The data

In most of the past posts, I used data from R or package. But this time is a bit special: the dataset is from me!

I've been recording the number of black kite at the bridge from Nov. 2022 to Aug. 2023, pretty much every day except some times when I went out of town.

This is not the main focus of the post, but let's create a map of the site (which is often for ecological research):

```{r, fig.width = 5, fig.height = 4}
library(tidyverse)



```


## The time series plot

Break it down into various steps


```{r, fig.width = 5, fig.height = 4}
library(tidyverse)



```







## Summary



Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!


