---
title: 'Post #42. Accelerate your ggplots with the extension package "ggforce"'
description: |
  Come take a look at some cool "ggforce" functions that can help strengthen your ggplot visualization!
author:
  - name: Gen-Chang Hsu
date: 2025-06-28
output:
  distill::distill_article:
    self_contained: false
categories:
  - "2025"
preview: ../../homepage_images/Post42.png
draft: true
---

<!-- Webpage style -->

<style>
d-article a {
color: #2780e3 !important;
border-bottom: none !important;
}

d-article a:hover {
color: #2780e3 !important;
border-bottom: 2px solid !important;
}

d-byline {
margin-left: -10% !important;
}

d-title {
margin-left: -10%;
}

d-title h1{
font-size: 45px;
width: 130%;
}

d-article {
margin-left: -10%;
}

p {
width: 120%;
}

d-article h2 {
width: 120%;
}

d-article h3 {
width: 120%;
font-size: 28px;
}

div.article-footer{
width: 120%;
}

d-article div.sourceCode { 
width: 120% !important;
}

d-article pre { 
width: 120% !important;
}

div.l-body {
width: 120% !important;
}

d-article table thead tr {
height: 60px !important;
width: 210px !important;
}

d-article table thead th {
height: 60px !important;
font-size: 22px !important;
width: 210px !important;
}

d-article table tbody tr td {
font-size: 21px !important;
height: 60px !important;
text-align: center !important;
width: 210px !important;
}

div.l-body img {
margin-right: 30% !important;
}

</style>

<head>
<base target="_blank">
</head>

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F,
                      fig.align = "center")

```

<!-- Start of the article -->

## Introduction

The extension package [`ggforce`](https://ggforce.data-imaginist.com/index.html) provides miscellaneous functions for enhancing ggplot visualization. In particular, it has several functions that allow users to customize basic geometric elements such as lines and shapes, which may otherwise require more effort to achieve using the original ggplot functions. There are also some other interesting functions we'll explore. Hopefully, after reading this quick post, you'll be able to put these functions into use in the future!


## `ggforce` functionality

### 1. Functions for lines

We're going to check out three line functions:

<span style="display: block; margin-top: 0px; margin-left: 10px">(1) `geom_link()`, which is similar to `geom_segment()` but allows users to draw lines with color/alpha transparency/size gradients by mapping the computed variable "index" (values from 0 to 1 representing the progression of interpolated points along the line segment) to those aesthetics.</span>

<span style="display: block; margin-top: 5px; margin-left: 10px">(2) `geom_arc()`, which draws arcs based on a center point (origin), a radius, a start angle, and an end angle (both in radians).</span>

<span style="display: block; margin-top: 5px; margin-left: 10px">(3) `geom_diagonal()`, which draws sigmoidal diagonal curves between two endpoints (as opposed to straight lines by `geom_link()`).</span>

Let's see how these functions work:

```{r, fig.width = 4, fig.height = 4}
library(tidyverse)
library(ggforce)

### (1) geom_link()
# create a dataframe with the x- and y-coordinates of the segment endpoints
geom_link_dat <- data.frame(x = 1, y = 1, xend = 2, yend = 2)

# draw the segment with color and size gradients
ggplot(geom_link_dat) + 
  geom_link(aes(x = x, y = y, xend = xend, yend = yend,
                color = after_stat(index),  # map "index" to the color aesthetic to create a color gradient along the segment
                linewidth = after_stat(index)^2 - after_stat(index)), # map "index" to the linewidth aesthetic to create a size gradient along the segment
            n = 100,  # n controls the number of interpolated points between the two endpoints
            show.legend = F) +  
  scale_color_viridis_c() +
  scale_linewidth_continuous(range = c(1, 4)) + 
  theme_bw(base_size = 13) + 
  coord_fixed()

### (2) geom_arc()
# create a dataframe with the origin, radius, start angle, and end angle of the arc
geom_arc_dat <- data.frame(x0 = 0, y0 = 0, r = 1, start = 0.2 * pi, end = 1.8 * pi)

# draw the arc with color and size gradients
ggplot(geom_arc_dat) + 
  geom_arc(aes(x0 = x0, y0 = y0, r = r, start = start, end = end,
               color = after_stat(index),
               linewidth = after_stat(index)),
           lineend = "round",
           n = 100,
           show.legend = F) +  
  scale_color_distiller(palette = "Spectral") +
  scale_linewidth_continuous(range = c(1, 4)) + 
  theme_bw(base_size = 13) + 
  coord_fixed()

### (3) geom_diagonal
# create a dataframe with the x- and y-coordinates of the diagonal curve endpoints
geom_diagonal_dat <- data.frame(x = 1, y = 1, xend = 2, yend = 2)

# draw the diagonal curve with a color gradient
ggplot(geom_diagonal_dat) + 
  geom_diagonal(aes(x = x, y = y, xend = xend, yend = yend,
               color = after_stat(index)),
               strength = 2,  # strength controls the curvature/steepness of the diagonal line
               linewidth = 3,
           n = 100,
           show.legend = F) +  
    geom_diagonal(aes(x = x, y = y, xend = xend, yend = yend,
               color = after_stat(index)),
               strength = 1,
               linewidth = 2,
           n = 100,
           show.legend = F) +  
    geom_diagonal(aes(x = x, y = y, xend = xend, yend = yend,
               color = after_stat(index)),
               strength = 0.5,
               linewidth = 1,
           n = 100,
           show.legend = F) +  
  scale_color_distiller(palette = "Set1") +
  theme_bw(base_size = 13) + 
  coord_fixed()

```

### 2. Functions for shapes

Again, we have three shape functions to check out:

<span style="display: block; margin-top: 0px; margin-left: 10px">(1) `geom_regon()`, which draws regular polygons with any number of sides and any rotation angle.</span>

<span style="display: block; margin-top: 5px; margin-left: 10px">(2) `geom_arc_bar()`, which is similar to `geom_arc()` but draws sectors (pies) and arc bars instead of arc lines.</span>

<span style="display: block; margin-top: 5px; margin-left: 10px">(3) `geom_diagonal_wide()`, which is similar to `geom_diagonal()` but draws polygons defined by an upper and a lower diagonal curve.</span>


```{r, fig.width = 4, fig.height = 4}
### (1) geom_regon()
# create a dataframe with the origins of the polygons, numbers of sides, rotation angles, and polygon sizes (r)
geom_regon_dat <- data.frame(x0 = rep(1:3, 3), 
                             y0 = rep(3:1, each = 3),
                             sides = 3:11,
                             angle = seq(0, 160, by = 20),
                             r = rep(0.2, 9))  # r is the ratio of polygon size to plot area  

# draw the regular polygons
ggplot(geom_regon_dat) + 
  geom_regon(aes(x0 = x0, y0 = y0, sides = sides, angle = angle, r = r, fill = sides), show.legend = F) +  
  scale_fill_viridis_c() +
  theme_bw(base_size = 13) +
  coord_fixed()

### (2) geom_arc_bar()
# create a dataframe with the origins, start radii, end radii, start angles, and end angles of the arc bars
geom_arc_bar_dat <- data.frame(x0 = rep(0, 6), 
                               y0 = rep(0, 6), 
                               r0 = seq(0, 1.8, length.out = 6),  # the start radii
                               r = rep(2, 6),  # the end radii
                               start = c(0/3, 1/3, 2/3, 3/3, 4/3, 5/3) * pi,  # the start angles 
                               end = c(1/3, 2/3, 3/3, 4/3, 5/3, 6/3) * pi,  # the end angles
                               fill = letters[1:6])

# draw the sector and arc bars
ggplot(geom_arc_bar_dat) + 
  geom_arc_bar(aes(x0 = x0, y0 = y0, r0 = r0, r = r, start = start, end = end, fill = fill), show.legend = F) +  
  scale_fill_brewer(palette = "Set2") +
  theme_bw(base_size = 13) + 
  coord_fixed()

### (3) geom_diagonal_wide()
# create a dataframe with an upper and a lower diagonal curve
geom_diagonal_wide_dat <- data.frame(x = c(1, 2.5, 1, 2.5),
                                     y = c(1, 2, 1.5, 2.5),
                                     group = c(1, 1, 1, 1))  # "group" identifies the four corner points defining each polygon   

# draw the diagonal polygon
ggplot(geom_diagonal_wide_dat) +
  geom_diagonal_wide(aes(x = x, y = y, group = group), 
                     strength = 0.7,  # steepness of the upper and lower diagonal curves 
                     radius = unit(0.2, "cm")) +  # rounded corners
  theme_bw(base_size = 13) + 
  coord_fixed()

```

### 3. Other interesting functions

We're going to take a look at a few other interesting (and potentially useful) `ggforce` functions:

<span style="display: block; margin-top: 0px; margin-left: 10px">(1) `geom_arc_bar(stats = "pie")`, which creates pie charts and donut charts with exploded slices. It works almost the same as `geom_arc_bar()`, but instead of start and end angles, we map the relative sizes of the slices to the "amount" aesthetic. Additionally, we can explode the slices using the "explode" aesthetic.</span>

```{r, fig.width = 4, fig.height = 4}
### (1) geom_arc_bar(stats = "pie")
# create a dataframe with the origins, start radii, end radii, relative sizes of the slices (amount), and degrees of slice explosion (explode)
geom_arc_bar_pie_dat <- data.frame(x0 = rep(0, 5), 
                                   y0 = rep(0, 5), 
                                   r0_pie = rep(0, 5),
                                   r0_donut = rep(1.3, 5),
                                   r = rep(2, 5),
                                   amount = c(1, 2, 3, 4, 5),
                                   explode = c(0, 0, 0, 0, 0.3))

# create a pie chart with an exploded slice
ggplot(geom_arc_bar_pie_dat) + 
  geom_arc_bar(aes(x0 = x0, y0 = y0, r0 = r0_pie, r = r, amount = amount, explode = explode, fill = as.factor(amount)), stat = "pie", show.legend = F) +  
  scale_fill_brewer(palette = "Dark2") +
  theme_bw(base_size = 13) + 
  coord_fixed()

### Create a donut chart
ggplot(geom_arc_bar_pie_dat) + 
  geom_arc_bar(aes(x0 = x0, y0 = y0, r0 = r0_donut, r = r, amount = amount, fill = as.factor(amount)), stat = "pie", show.legend = F) +  
  scale_fill_brewer(palette = "Dark2") +
  theme_bw(base_size = 13) + 
  coord_fixed()

```




geom_parallel_sets()

geom_voronoi()

geom_delaunay()

position_jitternormal()





## Summary


Hope you learn something useful from this post and don't forget to leave your comments and suggestions below if you have any!



